---

## ARIMA Forecasting: Temporal Inequality in Vaccination Progress

```{r setup, message=FALSE, warning=FALSE}
library(forecast)
library(ggplot2)
library(dplyr)
library(countrycode)


teal_dark   <- "#cc5500"
teal_mid    <- "#ff6200"
teal_base   <- "#ff8e00"
teal_light  <- "#ffb347"
teal_pale   <- "#ffd280"
teal_ghost  <- "#ffe8b8"


continent_colors <- c(
  "Africa"   = teal_dark,
  "Asia"     = teal_mid,
  "Oceania"  = teal_base,
  "Americas" = teal_light,
  "Europe"   = teal_pale
)


income_colors <- c(
  "High income"         = teal_dark,
  "Upper middle income" = teal_base,
  "Lower middle income" = teal_light,
  "Low income"          = teal_pale
)


theme_teal <- function() {
  theme_minimal() +
    theme(
      plot.title       = element_text(face = "bold", colour = "black"),
      plot.subtitle    = element_text(colour = "black"),
      axis.title       = element_text(colour = teal_dark),
      axis.text        = element_text(colour = "#555555"),
      legend.title     = element_text(colour = teal_dark, face = "bold"),
      panel.grid.major = element_line(colour = teal_ghost),
      panel.grid.minor = element_blank()
    )
}
```




---

## Germany: ARIMA Model

```{r germany_ts}
germany <- vaccinations_country %>%
  filter(location == "Germany",
         !is.na(people_fully_vaccinated_per_hundred)) %>%
  arrange(date)

ts_germany <- ts(
  germany$people_fully_vaccinated_per_hundred,
  frequency = 7
)
```

```{r germany_arima, message=FALSE, warning=FALSE}
fit <- auto.arima(ts_germany, d = 1, D = 1,
                  stepwise = FALSE, approximation = FALSE)
summary(fit)
```
```{r}
# Custom forecast plot function — full teal control
plot_forecast_teal <- function(fc, title_text, y_label) {
  
  # Extract components from the forecast object
  fitted_vals <- as.numeric(fc$x)
  forecast_vals <- as.numeric(fc$mean)
  
  n_fitted  <- length(fitted_vals)
  n_fore    <- length(forecast_vals)
  
  fitted_df <- data.frame(
    t     = seq_len(n_fitted),
    value = fitted_vals,
    type  = "Observed"
  )
  
  fore_df <- data.frame(
    t      = n_fitted + seq_len(n_fore),
    mean   = forecast_vals,
    lo80   = as.numeric(fc$lower[, 1]),
    hi80   = as.numeric(fc$upper[, 1]),
    lo95   = as.numeric(fc$lower[, 2]),
    hi95   = as.numeric(fc$upper[, 2])
  )
  
  ggplot() +
    # 95% CI band (outermost, palest)
    geom_ribbon(data = fore_df,
                aes(x = t, ymin = lo95, ymax = hi95),
                fill = teal_pale, alpha = 0.6) +
    # 80% CI band (inner, slightly darker)
    geom_ribbon(data = fore_df,
                aes(x = t, ymin = lo80, ymax = hi80),
                fill = teal_light, alpha = 0.7) +
    # Historical observed line
    geom_line(data = fitted_df,
              aes(x = t, y = value),
              colour = "#555555", linewidth = 0.7) +
    # Forecast mean line
    geom_line(data = fore_df,
              aes(x = t, y = mean),
              colour = teal_base, linewidth = 1) +
    labs(
      title    = title_text,
      subtitle = "30-day forecast with 80% and 95% confidence intervals",
      x        = "Time (weeks)",
      y        = y_label
    ) +
    theme_teal()
}
```

```{r germany_forecast}
forecast_30 <- forecast(fit, h = 30)

plot_forecast_teal(
  forecast_30,
  title_text = "ARIMA Forecast: Germany Full Vaccination Coverage",
  y_label    = "Fully vaccinated per 100"
) +
  scale_fill_manual(values  = c("80%" = teal_light,
                                "95%" = teal_pale)) +
  scale_colour_manual(values = c("Point Forecast" = red_base)) +
  labs(
    title    = "ARIMA Forecast: Germany Full Vaccination Coverage",
    subtitle = "30-day ahead forecast with 80% and 95% confidence intervals",
    x        = "Time (weeks)",
    y        = "Fully vaccinated per 100"
  ) +
  theme_teal()
```

---

## Residual Diagnostics

```{r residuals}
checkresiduals(fit)
```

The residual diagnostics for the ARIMA(2,1,2)(0,1,1)[7] model indicate a reasonable fit for Germany's vaccination data. The residual time series fluctuates randomly around zero with no systematic trend, suggesting the model captured the main dynamics of the series. A small number of larger residuals around observations 18–22 likely reflect irregular reporting patterns rather than model misspecification. The ACF plot shows no significant autocorrelation at most lags, with minor exceptions near lags 7 and 14, consistent with the weekly reporting structure of the data. The residual distribution approximates normality, supporting the validity of the confidence intervals produced by the forecast.

---

## ETS Model Comparison

```{r ets_forecast}
fit_ets <- ets(ts_germany)
fc_ets  <- forecast(fit_ets, h = 30)

plot_forecast_teal(
  fc_ets,
  title_text = "ETS Forecast: Germany Full Vaccination Coverage",
  y_label    = "Fully vaccinated per 100"
)
```

---

## ARIMA vs ETS Accuracy (Train/Test Split)

```{r accuracy_comparison, message=FALSE, warning=FALSE}
n     <- nrow(germany)
train <- germany$people_fully_vaccinated_per_hundred[1:round(n * 0.8)]
test  <- germany$people_fully_vaccinated_per_hundred[(round(n * 0.8) + 1):n]

ts_train <- ts(train, frequency = 7)

fit_arima_train <- auto.arima(ts_train, d = 1, D = 1,
                               stepwise = FALSE, approximation = FALSE)
fit_ets_train   <- ets(ts_train)

h <- length(test)

acc_arima <- accuracy(forecast(fit_arima_train, h = h), test)
acc_ets   <- accuracy(forecast(fit_ets_train,   h = h), test)

acc_arima
acc_ets
```

---

## Multi-Country ARIMA Forecast (Inequality Comparison)

```{r multicountry_arima, message=FALSE, warning=FALSE}
countries_arima <- c("Germany", "India", "Bangladesh")

for (ctry in countries_arima) {

  df <- vaccinations_country %>%
    filter(location == ctry,
           !is.na(people_fully_vaccinated_per_hundred)) %>%
    arrange(date) %>%
    mutate(
      daily_change = people_fully_vaccinated_per_hundred -
                     lag(people_fully_vaccinated_per_hundred)
    ) %>%
    filter(!is.na(daily_change))

  if (nrow(df) < 14) {
    message(paste("Skipping", ctry, "— not enough data"))
    next
  }

  ts_ctry <- ts(df$daily_change, frequency = 7)

  fit_ctry <- auto.arima(ts_ctry,
                          d = 1, D = 1,
                          stepwise      = FALSE,
                          approximation = FALSE)

  fc_ctry <- forecast(fit_ctry, h = 30)


  print(
    plot_forecast_teal(
      fc_ctry,
      title_text = paste("ARIMA Forecast of Daily Vaccination Change:", ctry),
      y_label    = "Daily change in fully vaccinated per 100"
    )
   +
      scale_fill_manual(values  = c("80%" = teal_light,
                                    "95%" = teal_pale)) +
      scale_colour_manual(values = c("Point Forecast" = teal_base)) +
      labs(
        title    = paste("ARIMA Forecast of Daily Vaccination Change:", ctry),
        subtitle = "30-day forecast with 80% and 95% confidence intervals",
        x        = "Time (weeks)",
        y        = "Daily change in fully vaccinated per 100"
      ) +
      theme_teal()
  )
}
```

The ARIMA forecast for Germany shows a declining trend in daily 
vaccination changes over the 30-day forecast horizon. The confidence 
intervals are relatively narrow, reflecting the model's high certainty 
due to Germany's consistent and well-reported vaccination data across 
244 observations. The predicted daily increments converge toward zero, 
consistent with a mature vaccination campaign approaching saturation 
in coverage.

The ARIMA forecast for India shows greater uncertainty than Germany, 
reflected in wider confidence intervals. The historical series reveals 
several sharp spikes in daily vaccination changes, likely corresponding 
to large-scale vaccination drives and reporting irregularities. Despite 
186 observations, the irregular pattern makes future vaccination 
increments harder to predict. The forecast suggests daily changes 
stabilising at a low positive level, indicating a slowing but ongoing 
vaccination campaign.

The ARIMA forecast for Bangladesh exhibits the widest confidence 
intervals of the three countries, reflecting the highest level of 
uncertainty. The historical series shows extended flat periods 
interrupted by sudden spikes, consistent with batch reporting rather 
than continuous data collection. With only 93 observations, the model 
has less data to learn from, resulting in an oscillating forecast with 
a broad uncertainty range. This instability illustrates a key dimension 
of global health inequality: lower-income countries not only vaccinated 
more slowly, but their inconsistent data infrastructure makes reliable 
forecasting — and therefore effective public health planning — 
substantially more difficult.


Comparing the three ARIMA forecasts reveals a clear pattern aligned 
with the broader inequality findings of this study. Germany, a 
high-income country, produced the most reliable forecast with narrow 
confidence intervals and a smooth historical series. India and 
Bangladesh, both lower-middle-income countries, showed progressively 
wider uncertainty bands and more irregular historical patterns. 
This gradient in forecast reliability mirrors the gradient in 
vaccination coverage and rollout speed observed earlier, suggesting 
that global health inequality manifests not only in vaccination outcomes 
but also in the quality and consistency of health data reporting.


