---
title: "COVID-19 Vaccination Trends and Global Health Inequality"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---
```{r}
loadings <- loadings %>%
  mutate(variable = recode(variable,
    "people_fully_vaccinated_per_hundred" = "Fully vaccinated",
    "total_vaccinations_per_hundred"      = "Total doses",
    "people_vaccinated_per_hundred"       = "At least one dose",
    "daily_vaccinations_per_million"      = "Daily doses / million"
  ))
```


```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(forcats)
library(tidyr)
library(scales)
library(factoextra)
library(corrplot)
library(countrycode)
library(ggrepel)

Sys.setlocale("LC_TIME", "C")

# ── Orange colour palette ──────────────────────────────────────────────────
orange_dark   <- "#cc5500"
orange_mid    <- "#ff6200"
orange_base   <- "#ff8e00"
orange_light  <- "#ffb347"
orange_pale   <- "#ffd280"
orange_ghost  <- "#ffe8b8"

# Named palettes for multi-category plots
country_colors <- c(
  "Germany"       = orange_dark,
  "Spain"         = orange_mid,
  "Czechia"       = orange_base,
  "Brazil"        = orange_light,
  "India"         = orange_pale,
  "United States" = "#e8a000"
)

continent_colors <- c(
  "Africa"   = orange_dark,
  "Asia"     = orange_mid,
  "Oceania"  = orange_base,
  "Americas" = orange_light,
  "Europe"   = orange_pale
)

income_colors <- c(
  "High income"         = orange_dark,
  "Upper middle income" = orange_base,
  "Lower middle income" = orange_light,
  "Low income"          = orange_pale
)

cluster_colors <- c(
  "1" = orange_dark,
  "2" = orange_base,
  "3" = orange_pale
)

# Shared ggplot2 theme
theme_orange <- function() {
  theme_minimal() +
    theme(
      plot.title       = element_text(face = "bold", colour = "black"),
      plot.subtitle    = element_text(colour = "black"),
      axis.title       = element_text(colour = orange_dark),
      axis.text        = element_text(colour = "#555555"),
      legend.title     = element_text(colour = orange_dark, face = "bold"),
      legend.text      = element_text(colour = "#555555"),
      panel.grid.major = element_line(colour = orange_ghost),
      panel.grid.minor = element_blank()
    )
}
```


```{r load_data}
library(dplyr)
library(stringr)
library(lubridate)

vaccinations <- read.csv("vaccinations.csv", sep = ",")


clean_percent_column <- function(x) {
  x %>%
    str_trim() %>%
    
    # roman numerals at beginning
    str_replace_all("^IV\\.", "4.") %>%
    str_replace_all("^III\\.", "3.") %>%
    str_replace_all("^II\\.", "2.") %>%
    str_replace_all("^I\\.", "1.") %>%
    
    # roman numerals at end
    str_replace_all("\\.VII$", ".7") %>%
    str_replace_all("\\.VI$", ".6") %>%
    str_replace_all("\\.V$", ".5") %>%
    str_replace_all("\\.IV$", ".4") %>%
    str_replace_all("\\.III$", ".3") %>%
    str_replace_all("\\.II$", ".2") %>%
    str_replace_all("\\.I$", ".1") %>%
    
    str_replace("^0+", "") %>%
    na_if("") %>%
    as.numeric()
}



vaccinations_clean <- vaccinations %>%
  mutate(
    date = as.Date(date),
    total_vaccinations_per_hundred =
      clean_percent_column(total_vaccinations_per_hundred),
    people_vaccinated_per_hundred =
      clean_percent_column(people_vaccinated_per_hundred),
    people_fully_vaccinated_per_hundred =
      clean_percent_column(people_fully_vaccinated_per_hundred),
    total_boosters_per_hundred =
      clean_percent_column(total_boosters_per_hundred)
  )
```

```{r data_prep}
regional_aggregates <- c(
  "World", "Europe", "European Union", "Asia", "Africa",
  "North America", "South America", "Oceania",
  "High income", "Upper middle income",
  "Lower middle income", "Low income"
)

vaccinations_country <- vaccinations_clean %>%
  filter(!location %in% regional_aggregates)

latest_country <- vaccinations_country %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  group_by(location) %>%
  slice_max(date, n = 1) %>%
  ungroup()

# ← PASTE HERE
missing_country <- vaccinations_country %>%
  group_by(location) %>%
  summarise(
    missing_pct = mean(is.na(people_fully_vaccinated_per_hundred)) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(missing_pct))
```

```{r}
latest_country <- latest_country %>%
  mutate(
    location = recode(
      location,
      "England" = "United Kingdom",
      "Scotland" = "United Kingdom",
      "Wales" = "United Kingdom",
      "Timor" = "Timor-Leste"
    )
  )
```


```{r missing_data}
missing_country <- vaccinations_country %>%
  group_by(location) %>%
  summarise(
    missing_pct = mean(is.na(people_fully_vaccinated_per_hundred)) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(missing_pct))
```

---

## Figure 1: Vaccination Trends in Selected Countries


```{r fig1_selected_countries}
selected_countries <- c(
  "Spain", "Germany", "Czechia", "Brazil", "India"
)

plot_data <- vaccinations_country %>%
  filter(
    location %in% selected_countries,
    !is.na(people_fully_vaccinated_per_hundred)
  )

last_points <- plot_data %>%
  group_by(location) %>%
  slice_max(date, n = 1) %>%
  ungroup()

ggplot(
  plot_data,
  aes(
    x = date,
    y = people_fully_vaccinated_per_hundred,
    colour = location
  )
) +
  geom_line(linewidth = 1.2) +

  geom_point(
    data = last_points,
    size = 3
  ) +

  geom_text_repel(
    data = last_points,
    aes(label = location),
    direction = "y",
    hjust = 0,
    nudge_x = 8,
    size = 3.8,
    fontface = "bold",
    segment.color = "grey70",
    show.legend = FALSE
  ) +

  geom_hline(
    yintercept = 50,
    linetype = "dashed",
    colour = "grey40",
    linewidth = 0.6
  ) +

  scale_color_manual(values = country_colors) +

  scale_x_date(
    date_labels = "%b %Y",
    expand = expansion(mult = c(0.02, 0.18))
  ) +

  labs(
    title = "COVID-19 Full Vaccination Trends in Selected Countries",
    subtitle = "Share of people fully vaccinated per 100 inhabitants",
    x = "Date",
    y = "Fully vaccinated per 100",
    caption = "Dashed line = 50% vaccination threshold"
  ) +

  theme_orange() +

  theme(
    legend.position = "none",
    plot.caption = element_text(colour = "grey40")
  )
```


Figure 1 illustrates substantial differences in vaccination rollout trajectories across countries. Spain, Germany and Czechia show faster growth than Brazil and India in the selected comparison. India remains clearly lower during the observed period, while Brazil increases more gradually. These patterns suggest unequal access to vaccines and differences in healthcare capacity and vaccination infrastructure.

---

## Figure 2 & 3: Top and Bottom 5 Countries

```{r fig2_top10}
top10 <- latest_country %>%
  arrange(desc(people_fully_vaccinated_per_hundred)) %>%
  slice_head(n = 10)

ggplot(
  top10,
  aes(
    x = reorder(location, people_fully_vaccinated_per_hundred),
    y = people_fully_vaccinated_per_hundred
  )
) +
  geom_col(
    fill = orange_base,
    width = 0.65,
    colour = "white",
    linewidth = 0.3
  ) +
  geom_text(
    aes(
      label = sprintf("%.1f%%", people_fully_vaccinated_per_hundred)
    ),
    hjust = -0.15,
    size = 3.5,
    colour = orange_dark,
    fontface = "bold"
  ) +
  coord_flip() +
  expand_limits(y = max(top10$people_fully_vaccinated_per_hundred) * 1.08) +
  labs(
    title    = "Top 10 Countries by Vaccination Coverage",
    subtitle = "Latest available observation",
    x        = "Country",
    y        = "Fully vaccinated per 100 inhabitants"
  ) +
  theme_orange()
```

```{r fig3_bottom10}
bottom10 <- latest_country %>%
  arrange(people_fully_vaccinated_per_hundred) %>%
  slice_head(n = 10)

ggplot(
  bottom10,
  aes(
    x = reorder(location, people_fully_vaccinated_per_hundred),
    y = people_fully_vaccinated_per_hundred
  )
) +
  geom_col(
    fill = orange_dark,
    width = 0.65,
    colour = "white",
    linewidth = 0.3
  ) +
  geom_text(
    aes(
      label = sprintf("%.2f", people_fully_vaccinated_per_hundred)
    ),
    hjust = -0.15,
    size = 3.5,
    colour = orange_dark,
    fontface = "bold"
  ) +
  coord_flip() +
  expand_limits(
    y = max(bottom10$people_fully_vaccinated_per_hundred) * 1.15
  ) +
  labs(
    title    = "Bottom 10 Countries by Vaccination Coverage",
    subtitle = "Latest available observation",
    x        = "Country",
    y        = "Fully vaccinated per 100 inhabitants"
  ) +
  theme_orange()
```

Figures 2 and 3 demonstrate substantial differences in vaccination coverage across countries. While the most vaccinated countries exceeded 70 fully vaccinated individuals per 100 inhabitants by August 2021, several countries remained below 1%. This contrast highlights the unequal distribution of vaccines and the differing capacities of national healthcare systems during the early stages of the global vaccination campaign.

---

## Speed of Vaccination Rollout

```{r rollout_speed}
rollout_speed <- vaccinations_country %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  group_by(location) %>%
  summarise(
    first_vaccination_date = min(date),
    reached_50 = any(people_fully_vaccinated_per_hundred >= 50),
    date_50 = ifelse(
      reached_50,
      min(date[people_fully_vaccinated_per_hundred >= 50]),
      NA
    ),
    date_50    = as.Date(date_50, origin = "1970-01-01"),
    days_to_50 = as.numeric(date_50 - first_vaccination_date),
    final_coverage = max(people_fully_vaccinated_per_hundred, na.rm = TRUE),
    .groups = "drop"
  )

rollout_speed %>% count(reached_50)
```

To evaluate differences in vaccination rollout speed, countries were classified according to whether they reached 50 fully vaccinated individuals per 100 inhabitants during the study period. The analysis revealed that only 61 out of 216 countries (28.2%) reached the 50% vaccination threshold by August 2021. In contrast, 155 countries (71.8%) failed to reach this level. These findings indicate substantial inequalities in vaccine access and rollout capacity.

---

## Figure 4: Fastest Countries to 50% Coverage

```{r fig4_fastest}
fastest10 <- rollout_speed %>%
  filter(reached_50) %>%
  arrange(days_to_50) %>%
  slice_head(n = 10)

ggplot(
  fastest10,
  aes(
    x = reorder(location, days_to_50),
    y = days_to_50
  )
) +
  geom_col(
    fill = orange_base,
    width = 0.65,
    colour = "white",
    linewidth = 0.3
  ) +
  geom_text(
    aes(label = days_to_50),
    hjust = -0.15,
    size = 3.5,
    colour = orange_dark,
    fontface = "bold"
  ) +
  coord_flip() +
  expand_limits(y = max(fastest10$days_to_50) * 1.08) +
  labs(
    title    = "Fastest Countries to Reach 50% Vaccination Coverage",
    subtitle = "Time from first vaccination record to the 50% threshold",
    x        = "Country",
    y        = "Days"
  ) +
  theme_orange()
```

---

## Figure 5: Distribution of Days to 50%

```{r fig5_days_histogram}
days_50_data <- rollout_speed %>%
  filter(reached_50 == TRUE)

median_days <- median(days_50_data$days_to_50, na.rm = TRUE)
mean_days   <- mean(days_50_data$days_to_50, na.rm = TRUE)

ggplot(days_50_data, aes(x = days_to_50)) +
  geom_histogram(
    bins = 15,
    fill = orange_light,
    colour = "white",
    alpha = 0.85
  ) +
  geom_vline(
    xintercept = median_days,
    linetype = "dashed",
    colour = orange_dark,
    linewidth = 1
  ) +
  geom_vline(
    xintercept = mean_days,
    linetype = "dotted",
    colour = orange_dark,
    linewidth = 1
  ) +
  annotate(
    "label",
    x = median_days + 5,
    y = 22,
    label = paste0(
      "Median = ", round(median_days), " days",
      "\nMean = ", round(mean_days), " days"
    ),
    fill = "white",
    colour = orange_dark,
    label.size = 0.2,
    size = 3.3
  ) +
  geom_rug(
    sides = "b",
    colour = orange_dark,
    alpha = 0.45
  ) +
  labs(
    title    = "Distribution of Days to Reach 50% Full Vaccination Coverage",
    subtitle = "Among countries that reached the threshold",
    x        = "Days to reach 50%",
    y        = "Number of countries"
  ) +
  theme_orange()
```

Among the 61 countries that reached 50% full vaccination coverage, the median time was 160 days. The mean (139 days) was lower, indicating a small number of countries achieved exceptionally rapid campaigns. Substantial variation — from a few weeks to 235 days — reflects differences in vaccine availability, healthcare infrastructure, and policy implementation.

---

## Figure 6: Lowest Vaccination Coverage

```{r fig6_worst20}
lowest10 <- rollout_speed %>%
  filter(!reached_50) %>%
  arrange(final_coverage) %>%
  slice_head(n = 10)

ggplot(
  lowest10,
  aes(
    x = reorder(location, final_coverage),
    y = final_coverage
  )
) +
  geom_col(
    fill = orange_dark,
    width = 0.65,
    colour = "white",
    linewidth = 0.3
  ) +
  geom_text(
    aes(label = sprintf("%.2f", final_coverage)),
    hjust = -0.15,
    size = 3.5,
    colour = orange_dark,
    fontface = "bold"
  ) +
  coord_flip() +
  expand_limits(y = max(lowest10$final_coverage) * 1.15) +
  labs(
    title    = "Lowest Vaccination Coverage by August 2021",
    subtitle = "Countries that remained furthest from the 50% threshold",
    x        = "Country",
    y        = "Fully vaccinated per 100 inhabitants"
  ) +
  theme_orange()
```

The Democratic Republic of Congo (0.01), Haiti (0.03), and South Sudan (0.04) reported the lowest vaccination coverage. The difference between the highest and lowest observed coverage exceeded a factor of 11,000 — strong evidence of unequal access to vaccines and healthcare resources.



---

## Figure 7: Missing Data by Country

```{r fig7_missing}
missing_country %>%
  filter(missing_pct >= 95) %>%
  slice_max(missing_pct, n = 15) %>%
  ggplot(aes(x = fct_reorder(location, missing_pct), y = missing_pct)) +
  geom_segment(aes(xend = location, y = 95, yend = missing_pct),
               colour = orange_dark) +
  geom_point(size = 3, colour = orange_dark) +
  coord_flip() +
  scale_y_continuous(limits = c(95, 100), breaks = seq(95, 100, 15)) +
  labs(
    title    = "Countries with Near-Complete Missing Vaccination Data",
    subtitle = "Five countries where 95% or more of records are missing",
    x        = "Country",
    y        = "Missing values (%)"
  ) +
  theme_orange()
```

The lollipop chart reveals that several countries show near-complete 
missingness in vaccination records, with Tokelau, Tuvalu, Madagascar, 
and others reaching 100% missing values. Importantly, high missingness 
does not necessarily indicate the absence of a vaccination campaign. 
Tokelau, for example, achieved over 96% vaccination coverage of its 
eligible population according to external sources (ReliefWeb, 2021), 
yet reported no data to the Our World in Data database. This discrepancy 
illustrates a critical limitation of the dataset: it captures reporting 
capacity rather than vaccination activity. Countries with limited 
administrative and digital infrastructure are systematically 
underrepresented, which means the inequalities observed in this analysis 
may in fact be underestimated. The absence of data is itself a signal 
of structural disadvantage.

---

## Figure 8: Vaccination Coverage by Income Group Over Time

```{r fig8_income_timeseries}

income_group_levels <- c(
  "High income",
  "Upper middle income",
  "Lower middle income",
  "Low income"
)

vaccinations_income <- vaccinations_clean %>%
  filter(location %in% income_group_levels) %>%
  mutate(
    location = factor(location, levels = income_group_levels)
  )

last_points <- vaccinations_income %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  group_by(location) %>%
  slice_max(date, n = 1) %>%
  ungroup()

ggplot(
  vaccinations_income %>%
    filter(!is.na(people_fully_vaccinated_per_hundred)),
  aes(
    x = date,
    y = people_fully_vaccinated_per_hundred,
    colour = location
  )
) +
  geom_line(linewidth = 1.2) +

  geom_point(
    data = last_points,
    size = 3
  ) +

  geom_text_repel(
    data = last_points,
    aes(
        label = paste0(
            location,
            " (",
            round(people_fully_vaccinated_per_hundred,1),
            ")"
        )
    ),
    direction = "y",
    hjust = 0,
    nudge_x = 8,
    size = 3.6,
    fontface = "bold",
    segment.color = "grey70",
    show.legend = FALSE
  ) +

  scale_color_manual(values = income_colors) +

  scale_x_date(
    date_labels = "%b %Y",
    expand = expansion(mult = c(0.02, 0.48))
  ) +

  labs(
    title = "COVID-19 Full Vaccination Coverage by Income Group",
    subtitle = "Aggregated data from Our World in Data",
    x = "Date",
    y = "Fully vaccinated per 100"
  ) +

  theme_orange() +

  theme(
    legend.position = "none"
  )
```

The income group time series reveals a widening gap throughout 2021. High-income countries accelerated rapidly while low-income countries remained near zero — providing direct evidence of structural global health inequality in vaccine distribution.

---

## Figure 9: Vaccination Coverage by Continent
```{r}
library(ggdist)
library(ggbeeswarm)
library(countrycode)
```

```{r add_continent, warning=FALSE}
latest_country <- latest_country %>%
  mutate(continent = countrycode(location,
                                 origin      = "country.name",
                                 destination = "continent"))
```

```{r fig9_continent_raincloud, warning=FALSE}
latest_country %>%
  filter(
    !is.na(continent),
    !is.na(people_fully_vaccinated_per_hundred),
    people_fully_vaccinated_per_hundred <= 100
  ) %>%
  mutate(
    continent = reorder(
      continent,
      people_fully_vaccinated_per_hundred,
      median
    )
  ) %>%
  ggplot(aes(
    x = continent,
    y = people_fully_vaccinated_per_hundred,
    fill = continent,
    color = continent
  )) +
  ggdist::stat_halfeye(
    adjust = 0.7,
    width = 0.8,
    justification = -0.18,
    .width = 0,
    point_colour = NA,
    alpha = 0.65
  ) +

  geom_boxplot(
    width = 0.16,
    outlier.shape = NA,
    alpha = 0.45,
    colour = orange_dark
  ) +
  ggbeeswarm::geom_quasirandom(
    width = 0.08,
    size = 1.5,
    alpha = 0.45,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = continent_colors) +
  scale_color_manual(values = continent_colors) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20)
  ) +
  labs(
    title    = "Vaccination Coverage by Continent",
    subtitle = "Latest available observation per country",
    x        = "Continent",
    y        = "Fully vaccinated per 100"
  ) +
  theme_orange() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 20, hjust = 1)
  )
```

Africa's median vaccination coverage was near zero, while Europe's median exceeded 50 fully vaccinated per 100 inhabitants. This continental disparity mirrors the income group findings and further confirms that geography and economic development were strongly associated with vaccination outcomes.

---

## Figure 10: Correlation Matrix

```{r fig10_correlation}
library(ggcorrplot)

correlation_data <- latest_country %>%
  select(
    total_vaccinations_per_hundred,
    people_vaccinated_per_hundred,
    people_fully_vaccinated_per_hundred,
    daily_vaccinations_per_million
  )

cor_matrix <- cor(correlation_data, use = "complete.obs")

ggcorrplot(
  cor_matrix,
  method = "square",
  type = "lower",
  lab = TRUE,
  lab_size = 4,
  colors = c(orange_ghost, "white", orange_dark),
  outline.col = "white",
  title = "Correlation Matrix of Vaccination Indicators"
) +
  theme_orange() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

Strong positive correlations (0.94–0.98) were observed between cumulative vaccination indicators. Daily vaccinations per million showed only weak to moderate correlations (0.16–0.40) with cumulative measures, suggesting that current vaccination speed did not necessarily reflect overall success — global differences were primarily driven by cumulative rollout capacity.

---

## Figure 11: PCA of Vaccination Indicators

```{r fig11_pca}
library(ggrepel)

label_countries <- c("Spain", "Germany", "Czechia", "Brazil", "India")

ggplot(pca_scores, aes(PC1, PC2)) +

  geom_point(
    aes(colour = continent),
    size = 2.5,
    alpha = 0.75
  ) +

  geom_point(
    data = subset(pca_scores, location %in% label_countries),
    shape = 21,
    fill = "white",
    colour = "black",
    stroke = 0.8,
    size = 3.8
  ) +

  geom_text_repel(
    data = subset(pca_scores, location %in% label_countries),
    aes(label = location),
    size = 3.6,
    fontface = "bold",
    box.padding = 0.5,
    point.padding = 0.3
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    colour = "grey70"
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    colour = "grey70"
  ) +

  scale_colour_manual(values = continent_colors) +

  labs(
    title = "PCA of Countries by Vaccination Profile",
    subtitle = "Countries coloured according to continent",
    x = "PC1 (overall vaccination coverage)",
    y = "PC2 (daily vaccination intensity)",
    colour = "Continent"
  ) +

  theme_orange() +
  theme(
    legend.position = "bottom"
  )
```
```{r}
unique(
  pca_scores$location[is.na(pca_scores$continent)]
)
```




```{r}
ggplot() +

  geom_point(
    data = pca_scores,
    aes(PC1, PC2),
    colour = "grey80",
    alpha = 0.35,
    size = 2
  ) +

  geom_segment(
    data = loadings,
    aes(
      x = 0,
      y = 0,
      xend = PC1,
      yend = PC2
    ),
    arrow = arrow(length = unit(0.3, "cm")),
    colour = orange_dark,
    linewidth = 1
  ) +

  geom_text_repel(
    data = loadings,
    aes(
      PC1,
      PC2,
      label = variable
    ),
    colour = orange_dark,
    fontface = "bold",
    size = 4,
    force = 10,
    max.overlaps = Inf,
    box.padding = 0.7
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    colour = "grey70"
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    colour = "grey70"
  ) +

  labs(
    title = "PCA Variable Loadings",
    subtitle = "Arrows indicate the contribution of vaccination indicators to the principal components",
    x = "PC1",
    y = "PC2"
  ) +

  theme_orange()
```
The PCA variable loadings plot confirms the structure of the first 
principal component. Three vaccination indicators — fully vaccinated 
per hundred, total doses per hundred, and at least one dose per hundred 
— load strongly and in the same direction along PC1, indicating that 
countries performing well on one cumulative metric tend to perform well 
on all of them. Daily vaccinations per million loads primarily on PC2 
and is nearly orthogonal to the cumulative indicators, confirming that 
current vaccination intensity was largely independent of overall 
campaign success. This supports the interpretation of PC1 as a summary 
axis of vaccination inequality.
---

## Figure 12: K-means Clustering in PCA Space


```{r fig12_kmeans}
# Build pca_input keeping location for clustering
pca_input <- latest_country %>%
  select(location,
         total_vaccinations_per_hundred,
         people_vaccinated_per_hundred,
         people_fully_vaccinated_per_hundred,
         daily_vaccinations_per_million) %>%
  na.omit()

pca_data <- pca_input %>% select(-location)
pca      <- prcomp(pca_data, scale. = TRUE)

set.seed(42)
km <- kmeans(scale(pca_data), centers = 3, nstart = 25)

pca_scores          <- as.data.frame(pca$x)
pca_scores$cluster  <- factor(km$cluster)
pca_scores$location <- pca_input$location

label_countries <- c("Spain", "Germany", "Czechia", "Brazil", "India")

cluster_colors_named <- c(
  "1" = "#E07000",
  "2" = "#1F3864",
  "3" = "#FFB347"
)

ggplot(pca_scores, aes(PC1, PC2, color = cluster)) +
  stat_ellipse(aes(fill = cluster),
               geom = "polygon", alpha = 0.08,
               level = 0.85, show.legend = FALSE) +
  geom_point(alpha = 0.75, size = 2.8) +
  geom_text_repel(
    data          = pca_scores %>% filter(location %in% label_countries),
    aes(label     = location),
    size          = 2.8, color = "black", fontface = "bold",
    max.overlaps  = 30, box.padding = 0.5, point.padding = 0.3,
    segment.color = "#AAAAAA", segment.size = 0.3,
    min.segment.length = 0.2
  ) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "#CCCCCC", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "#CCCCCC", linewidth = 0.4) +
  scale_color_manual(
    values = cluster_colors_named,
    labels = c(
      "1 — Upper-middle coverage  (Brazil)",
      "2 — High coverage  (Spain, Germany, Czechia)",
      "3 — Low coverage  (India)"
    )
  ) +
  scale_fill_manual(values = cluster_colors_named) +
  labs(
    title    = "K-means Clusters in PCA Space (k = 3)",
    subtitle = "Countries grouped by vaccination profile — selected countries labelled",
    x        = "PC1 (76.5% variance — overall vaccination success)",
    y        = "PC2 (22.8% variance — daily vaccination intensity)",
    color    = "Cluster",
    caption  = "Ellipses show 85% confidence region per cluster"
  ) +
  theme_orange() +
  theme(legend.text = element_text(size = 8.5),
        legend.key.size = unit(0.9, "lines"),
        plot.caption = element_text(size = 8, color = "#AAAAAA", hjust = 1))

```
K-means clustering relies on a random initialisation of cluster centroids, 
meaning that results may differ across repeated runs of the algorithm. 
To ensure reproducibility of the analysis, a random seed was set prior 
to model fitting (set.seed(42)). This guarantees that the cluster 
assignments reported here can be exactly reproduced by any reader 
running the same code, which is a standard requirement for transparent 
and scientifically verifiable data analysis (Peng, 2011).

```{r kmeans_summary}
pca_data_clustered <- pca_data
pca_data_clustered$cluster <- factor(km$cluster)

pca_data_clustered %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean, na.rm = TRUE))
```

K-means clustering identified three distinct country groups. Cluster 1 corresponds to countries with high cumulative vaccination coverage, Cluster 2 to countries with moderate progress, and Cluster 3 to countries with minimal vaccination coverage. The clear separation between clusters reinforces the finding that global health inequality in vaccination outcomes was substantial and not continuous.

```{r}
# Rebuild pca_data with location attached
pca_input <- latest_country %>%
  select(location,
         total_vaccinations_per_hundred,
         people_vaccinated_per_hundred,
         people_fully_vaccinated_per_hundred,
         daily_vaccinations_per_million) %>%
  na.omit()

pca_data    <- pca_input %>% select(-location)
pca         <- prcomp(pca_data, scale. = TRUE)

set.seed(42)
km          <- kmeans(scale(pca_data), centers = 3, nstart = 25)

pca_scores          <- as.data.frame(pca$x)
pca_scores$cluster  <- factor(km$cluster)
pca_scores$location <- pca_input$location  # now location is attached
```

```{r}
library(ggrepel)

# Remove this line — it overwrites the correct location you just set
# pca_scores$location <- pca_data_clustered$location  ← DELETE THIS

label_countries <- c("Spain", "Germany", "Czechia", "Brazil", "India")

ggplot(pca_scores, aes(PC1, PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_text_repel(
    data        = pca_scores %>% filter(location %in% label_countries),
    aes(label   = location),
    size        = 2.8,
    color       = "black",
    max.overlaps = 25,
    box.padding  = 0.4,
    segment.color = "#AAAAAA",
    segment.size  = 0.3
  ) +
  scale_color_manual(
    values = cluster_colors,
    labels = c(
      "1 — Upper-middle coverage\n   (e.g. Brazil, Japan, Israel)",
      "2 — High coverage\n   (e.g. Germany, UK, US)",
      "3 — Low coverage\n   (e.g. India, DRC, Haiti)"
    )
  ) +
  labs(
    title    = "K-means Clusters in PCA Space (k = 3)",
    subtitle = "Countries grouped by vaccination profile — selected countries labelled",
    x        = "PC1 (76.5% variance explained)",
    y        = "PC2 (22.8% variance explained)",
    color    = "Cluster"
  ) +
  theme_orange() +
  theme(legend.text = element_text(size = 8))
```


```{r}
library(ggrepel)

# Identify the mystery outlier at the top
pca_scores %>% arrange(desc(PC2)) %>% select(location, cluster, PC1, PC2) %>% head(3)

label_countries <- c("Spain", "Germany", "Czechia", "Brazil", "India")

# Better distinct colors — not all orange shades
cluster_colors_named <- c(
  "1" = "#E07000",   # burnt orange — upper-middle
  "2" = "#1F3864",   # navy — high coverage (your report navy)
  "3" = "#FFB347"    # pale orange — low coverage
)

ggplot(pca_scores, aes(PC1, PC2, color = cluster)) +
  # Cluster ellipses to show grouping clearly
  stat_ellipse(aes(fill = cluster), 
               geom = "polygon", alpha = 0.08,
               level = 0.85, show.legend = FALSE) +
  geom_point(alpha = 0.75, size = 2.8) +
  geom_text_repel(
    data         = pca_scores %>% filter(location %in% label_countries),
    aes(label    = location),
    size         = 2.8,
    color        = "black",
    fontface     = "bold",
    max.overlaps = 30,
    box.padding  = 0.5,
    point.padding = 0.3,
    segment.color = "#AAAAAA",
    segment.size  = 0.3,
    min.segment.length = 0.2
  ) +
  scale_color_manual(
    values = cluster_colors_named,
    labels = c(
      "1 — Upper-middle coverage  (Brazil)",
      "2 — High coverage  (Spain, Germany, Czechia)",
      "3 — Low coverage  (India)"
    )
  ) +
  scale_fill_manual(
    values = cluster_colors_named
  ) +
  # Reference lines at origin
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "#CCCCCC", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", 
             color = "#CCCCCC", linewidth = 0.4) +
  labs(
    title    = "K-means Clusters in PCA Space (k = 3)",
    subtitle = "Countries grouped by vaccination profile — selected countries labelled",
    x        = "PC1 (76.5% variance explained — overall vaccination success)",
    y        = "PC2 (22.8% variance explained — daily vaccination intensity)",
    color    = "Cluster",
    caption  = "Ellipses show 85% confidence region per cluster"
  ) +
  theme_orange() +
  theme(
    legend.text     = element_text(size = 8.5),
    legend.key.size = unit(0.9, "lines"),
    plot.caption    = element_text(size = 8, color = "#AAAAAA", hjust = 1)
  )
```



---

## Summary

```{r final_summary}
final_summary <- latest_country %>%
  select(location, date,
         people_fully_vaccinated_per_hundred) %>%
  left_join(
    rollout_speed %>% select(location, reached_50, days_to_50),
    by = "location"
  ) %>%
  left_join(missing_country, by = "location") %>%
  arrange(desc(people_fully_vaccinated_per_hundred))

head(final_summary, 20)
```

COVID-19 vaccination trends differed substantially across countries. High-income countries reached higher coverage earlier, while lower-income countries experienced delayed rollout. The 50% threshold analysis demonstrates that inequality was visible not only in final coverage levels but also in rollout speed. Missing data patterns further suggest that global inequality extends to health data reporting infrastructure — with consequences for public health planning and resource allocation.
