---
title: "COVID-19 Vaccination Trends and Global Health Inequality"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(forcats)
library(tidyr)
library(scales)
library(factoextra)
library(corrplot)
library(countrycode)
library(stringr)
library(ggrepel)

Sys.setlocale("LC_TIME", "C")

orange_dark   <- "#cc5500"
orange_mid    <- "#ff6200"
orange_base   <- "#ff8e00"
orange_light  <- "#ffb347"
orange_pale   <- "#ffd280"
orange_ghost  <- "#ffe8b8"

country_colors <- c(
  "Germany"       = orange_dark,
  "United States" = orange_mid,
  "Czechia"       = orange_base,
  "Brazil"        = orange_light,
  "India"         = orange_pale,
  "South Africa"  = "#e8a000"
)

continent_colors <- c(
  "Africa"   = orange_dark,
  "Asia"     = orange_mid,
  "Oceania"  = orange_base,
  "Americas" = orange_light,
  "Europe"   = orange_pale
)

income_colors <- c(
  "High income"         = orange_dark,
  "Upper middle income" = orange_base,
  "Lower middle income" = orange_light,
  "Low income"          = orange_pale
)

theme_orange <- function() {
  theme_minimal() +
    theme(
      plot.title       = element_text(face = "bold", colour = "black"),
      plot.subtitle    = element_text(colour = "black"),
      axis.title       = element_text(colour = orange_dark),
      axis.text        = element_text(colour = "#555555"),
      legend.title     = element_text(colour = orange_dark, face = "bold"),
      legend.text      = element_text(colour = "#555555"),
      panel.grid.major = element_line(colour = orange_ghost),
      panel.grid.minor = element_blank()
    )
}
```

```{r load_data}
vaccinations <- read.csv("vaccinations.csv", sep = ";")

clean_percent_column <- function(x) {
  x %>%
    str_trim() %>%
    str_replace_all("^IV\\.", "14.") %>%
    str_replace_all("^III\\.", "111.") %>%
    str_replace_all("^II\\.", "11.") %>%
    str_replace_all("^I\\.", "1.") %>%
    str_replace_all("\\.VII$", ".7") %>%
    str_replace_all("\\.VI$", ".6") %>%
    str_replace_all("\\.V$", ".5") %>%
    str_replace_all("\\.IV$", ".4") %>%
    str_replace_all("\\.III$", ".3") %>%
    str_replace_all("\\.II$", ".2") %>%
    str_replace_all("\\.I$", ".1") %>%
    str_replace("^0+", "") %>%
    na_if("") %>%
    as.numeric()
}

vaccinations_clean <- vaccinations %>%
  mutate(
    date = dmy(date),
    total_vaccinations_per_hundred         = clean_percent_column(total_vaccinations_per_hundred),
    people_vaccinated_per_hundred          = clean_percent_column(people_vaccinated_per_hundred),
    people_fully_vaccinated_per_hundred    = clean_percent_column(people_fully_vaccinated_per_hundred),
    total_boosters_per_hundred             = clean_percent_column(total_boosters_per_hundred)
  ) %>%
  mutate(
    people_fully_vaccinated_per_hundred = ifelse(
      people_fully_vaccinated_per_hundred > 130, NA,
      people_fully_vaccinated_per_hundred
    )
  )
```

---

## Data Preparation

```{r data_prep}
regional_aggregates <- c(
  "World", "Europe", "European Union", "Asia", "Africa",
  "North America", "South America", "Oceania",
  "High income", "Upper middle income",
  "Lower middle income", "Low income"
)

vaccinations_country <- vaccinations_clean %>%
  filter(!location %in% regional_aggregates)

latest_country <- vaccinations_country %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  group_by(location) %>%
  slice_max(date, n = 1) %>%
  ungroup()

missing_country <- vaccinations_country %>%
  group_by(location) %>%
  summarise(
    missing_pct = mean(is.na(people_fully_vaccinated_per_hundred)) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(missing_pct))
```

The dataset covers `r n_distinct(vaccinations_country$location)` countries and territories, spanning from `r min(vaccinations_country$date)` to `r max(vaccinations_country$date)`. The latest available vaccination observation per country was used for cross-sectional comparisons.

---

## Figure 1: Vaccination Trends in Selected Countries

```{r fig1_selected_countries}
selected_countries <- c(
  "Germany", "Czechia", "South Africa",
  "India", "Brazil", "United States"
)

vaccinations_country %>%
  filter(location %in% selected_countries,
         !is.na(people_fully_vaccinated_per_hundred)) %>%
  ggplot(aes(
    x     = date,
    y     = people_fully_vaccinated_per_hundred,
    color = location
  )) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 50, linetype = "dashed",
             color = "black", linewidth = 0.5) +
  scale_color_manual(values = country_colors) +
  scale_x_date(date_labels = "%b %Y") +
  labs(
    title    = "COVID-19 Full Vaccination Trends in Selected Countries",
    subtitle = "Share of people fully vaccinated per 100 inhabitants",
    x        = "Date",
    y        = "Fully vaccinated per 100",
    color    = "Country",
    caption  = "Dashed line = 50% threshold"
  ) +
  theme_orange()
```

Figure 1 illustrates substantial differences in vaccination rollout trajectories across countries. The United States achieved rapid vaccination coverage early in 2021, while Germany and Czechia accelerated later and eventually surpassed the United States. In contrast, India and South Africa experienced considerably slower growth throughout the observed period. These patterns suggest unequal access to vaccines and differences in healthcare capacity and vaccination infrastructure.

---

## Figure 2 & 3: Top and Bottom 10 Countries

```{r fig2_top10}
top10 <- latest_country %>%
  arrange(desc(people_fully_vaccinated_per_hundred)) %>%
  slice_head(n = 10)

ggplot(top10,
       aes(x = reorder(location, people_fully_vaccinated_per_hundred),
           y = people_fully_vaccinated_per_hundred)) +
  geom_col(fill = orange_base) +
  coord_flip() +
  labs(
    title    = "Top 10 Countries by Vaccination Coverage",
    subtitle = "Latest available observation",
    x        = "Country",
    y        = "Fully vaccinated per 100 inhabitants"
  ) +
  theme_orange()
```

```{r fig3_bottom10}
bottom10 <- latest_country %>%
  arrange(people_fully_vaccinated_per_hundred) %>%
  slice_head(n = 10)

ggplot(bottom10,
       aes(x = reorder(location, people_fully_vaccinated_per_hundred),
           y = people_fully_vaccinated_per_hundred)) +
  geom_col(fill = orange_dark) +
  coord_flip() +
  labs(
    title    = "Bottom 10 Countries by Vaccination Coverage",
    subtitle = "Latest available observation",
    x        = "Country",
    y        = "Fully vaccinated per 100 inhabitants"
  ) +
  theme_orange()
```

Figures 2 and 3 demonstrate substantial differences in vaccination coverage across countries. While the most vaccinated countries exceeded 70 fully vaccinated individuals per 100 inhabitants by August 2021, several countries remained below 1%. This contrast highlights the unequal distribution of vaccines and the differing capacities of national healthcare systems during the early stages of the global vaccination campaign.

---

## Speed of Vaccination Rollout

```{r rollout_speed}
rollout_speed <- vaccinations_country %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  group_by(location) %>%
  summarise(
    first_vaccination_date = min(date),
    reached_50 = any(people_fully_vaccinated_per_hundred >= 50),
    date_50 = ifelse(
      reached_50,
      min(date[people_fully_vaccinated_per_hundred >= 50]),
      NA
    ),
    date_50    = as.Date(date_50, origin = "1970-01-01"),
    days_to_50 = as.numeric(date_50 - first_vaccination_date),
    final_coverage = max(people_fully_vaccinated_per_hundred, na.rm = TRUE),
    .groups = "drop"
  )

rollout_speed %>% count(reached_50)
```

To evaluate differences in vaccination rollout speed, countries were classified according to whether they reached 50 fully vaccinated individuals per 100 inhabitants during the study period. The analysis revealed that only 61 out of 216 countries (28.2%) reached the 50% vaccination threshold by August 2021. In contrast, 155 countries (71.8%) failed to reach this level. These findings indicate substantial inequalities in vaccine access and rollout capacity.

---

## Figure 4: Fastest Countries to 50% Coverage

```{r fig4_fastest}
fastest_20 <- rollout_speed %>%
  filter(reached_50 == TRUE) %>%
  arrange(days_to_50) %>%
  slice_head(n = 20)

ggplot(fastest_20,
       aes(x = reorder(location, -days_to_50),
           y = days_to_50)) +
  geom_col(fill = orange_base) +
  coord_flip() +
  labs(
    title    = "Fastest Countries to Reach 50% Full Vaccination Coverage",
    subtitle = "Countries with the shortest time to reach the 50% threshold",
    x        = "Country",
    y        = "Days from first vaccination record to 50%"
  ) +
  theme_orange()
```

---

## Figure 5: Distribution of Days to 50%

```{r fig5_days_histogram}
rollout_speed %>%
  filter(reached_50 == TRUE) %>%
  ggplot(aes(x = days_to_50)) +
  geom_histogram(bins = 15, fill = orange_base,
                 colour = orange_dark) +
  labs(
    title    = "Distribution of Days to Reach 50% Full Vaccination Coverage",
    subtitle = "Among 61 countries that reached the threshold",
    x        = "Days to reach 50%",
    y        = "Number of countries"
  ) +
  theme_orange()

rollout_speed %>%
  filter(reached_50 == TRUE) %>%
  summarise(
    min_days    = min(days_to_50,    na.rm = TRUE),
    median_days = median(days_to_50, na.rm = TRUE),
    mean_days   = mean(days_to_50,   na.rm = TRUE),
    max_days    = max(days_to_50,    na.rm = TRUE)
  )
```

Among the 61 countries that reached 50% full vaccination coverage, the median time was 160 days. The mean (139 days) was lower, indicating a small number of countries achieved exceptionally rapid campaigns. Substantial variation — from a few weeks to 235 days — reflects differences in vaccine availability, healthcare infrastructure, and policy implementation.

---

## Figure 6: Lowest Vaccination Coverage

```{r fig6_worst20}
never_50 <- rollout_speed %>%
  filter(reached_50 == FALSE) %>%
  arrange(final_coverage)

ggplot(never_50 %>% slice_head(n = 20),
       aes(x = reorder(location, final_coverage),
           y = final_coverage)) +
  geom_col(fill = orange_dark) +
  coord_flip() +
  labs(
    title    = "Lowest Observed Vaccination Coverage by August 2021",
    subtitle = "Countries furthest from the 50% threshold",
    x        = "Country",
    y        = "Fully vaccinated per 100"
  ) +
  theme_orange()
```

The Democratic Republic of Congo (0.01), Haiti (0.03), and South Sudan (0.04) reported the lowest vaccination coverage. The difference between the highest and lowest observed coverage exceeded a factor of 11,000 — strong evidence of unequal access to vaccines and healthcare resources.

---

## Figure 7: Missing Data by Country

```{r fig7_missing}
missing_country %>%
  filter(missing_pct >= 95) %>%
  slice_max(missing_pct, n = 20) %>%
  ggplot(aes(x = fct_reorder(location, missing_pct), y = missing_pct)) +
  geom_segment(aes(xend = location, y = 95, yend = missing_pct),
               colour = orange_dark) +
  geom_point(size = 3, colour = orange_dark) +
  coord_flip() +
  scale_y_continuous(limits = c(95, 100), breaks = seq(95, 100, 5)) +
  labs(
    title    = "Countries with Near-Complete Missing Vaccination Data",
    subtitle = "Countries where 95% or more of records are missing",
    x        = "Country",
    y        = "Missing values (%)"
  ) +
  theme_orange()
```

The lollipop chart reveals that several countries show near-complete missingness in vaccination records, with Tokelau, Tuvalu, Madagascar, and others reaching 100% missing values. Importantly, high missingness does not necessarily indicate the absence of a vaccination campaign. Tokelau, for example, achieved over 96% vaccination coverage of its eligible population according to external sources (ReliefWeb, 2021), yet reported no data to the Our World in Data database. This discrepancy illustrates a critical limitation of the dataset: it captures reporting capacity rather than vaccination activity. Countries with limited administrative and digital infrastructure are systematically underrepresented, which means the inequalities observed in this analysis may in fact be underestimated. The absence of data is itself a signal of structural disadvantage.

---

## Figure 8: Vaccination Coverage by Income Group Over Time

```{r fig8_income_timeseries}
income_group_levels <- c("High income", "Upper middle income",
                         "Lower middle income", "Low income")

vaccinations_income <- vaccinations_clean %>%
  filter(location %in% income_group_levels) %>%
  mutate(location = factor(location, levels = income_group_levels))

vaccinations_income %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  ggplot(aes(x     = date,
             y     = people_fully_vaccinated_per_hundred,
             color = location)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = income_colors) +
  scale_x_date(date_labels = "%b %Y") +
  labs(
    title    = "COVID-19 Full Vaccination Coverage by Income Group",
    subtitle = "Aggregated data from Our World in Data",
    x        = "Date",
    y        = "Fully vaccinated per 100",
    color    = "Income Group"
  ) +
  theme_orange()
```

The income group time series reveals a widening gap throughout 2021. High-income countries accelerated rapidly while low-income countries remained near zero — providing direct evidence of structural global health inequality in vaccine distribution.

---

## Figure 9: Vaccination Coverage by Continent

```{r fig9_continent_boxplot, warning=FALSE}
latest_country <- latest_country %>%
  mutate(
    continent = countrycode(location,
                            origin      = "country.name",
                            destination = "continent")
  )

latest_country %>%
  filter(!is.na(continent)) %>%
  ggplot(aes(
    x    = reorder(continent,
                   people_fully_vaccinated_per_hundred, median),
    y    = people_fully_vaccinated_per_hundred,
    fill = continent
  )) +
  geom_boxplot(show.legend = FALSE, colour = orange_dark, alpha = 0.85) +
  scale_fill_manual(values = continent_colors) +
  labs(
    title    = "Vaccination Coverage by Continent",
    subtitle = "Latest available observation per country",
    x        = "Continent",
    y        = "Fully vaccinated per 100"
  ) +
  theme_orange()
```

Africa's median vaccination coverage was near zero, while Europe's median exceeded 50 fully vaccinated per 100 inhabitants. This continental disparity mirrors the income group findings and further confirms that geography and economic development were strongly associated with vaccination outcomes.

---

## Figure 10: Correlation Matrix

```{r fig10_correlation}
correlation_data <- latest_country %>%
  select(
    total_vaccinations_per_hundred,
    people_vaccinated_per_hundred,
    people_fully_vaccinated_per_hundred,
    daily_vaccinations_per_million
  )

cor_matrix <- cor(correlation_data, use = "complete.obs")

corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  addCoef.col = "black",
  col         = colorRampPalette(c(orange_ghost, "white", orange_dark))(200),
  tl.col      = "black",
  tl.srt      = 45
)
```

Strong positive correlations (0.94--0.98) were observed between cumulative vaccination indicators. Daily vaccinations per million showed only weak to moderate correlations (0.16--0.40) with cumulative measures, suggesting that current vaccination speed did not necessarily reflect overall success -- global differences were primarily driven by cumulative rollout capacity.

---

## Figure 11: PCA of Vaccination Indicators

```{r fig11_pca}
pca_input <- latest_country %>%
  select(location,
         total_vaccinations_per_hundred,
         people_vaccinated_per_hundred,
         people_fully_vaccinated_per_hundred,
         daily_vaccinations_per_million) %>%
  na.omit()

pca_data <- pca_input %>% select(-location)
pca      <- prcomp(pca_data, scale. = TRUE)
summary(pca)

fviz_pca_var(
  pca,
  repel         = TRUE,
  col.var       = "contrib",
  gradient.cols = c(orange_pale, orange_base, orange_dark)
)
```

The first principal component explained 76.5% of total variance, the second 22.8% -- together 99.3%. Differences between countries were largely driven by a single dimension representing overall vaccination success, supporting the conclusion that disparities were substantial and systematic rather than random.

---

## Figure 12: K-means Clustering in PCA Space

K-means clustering relies on a random initialisation of cluster centroids, meaning that results may differ across repeated runs of the algorithm. To ensure reproducibility of the analysis, a random seed was set prior to model fitting (set.seed(42)). This guarantees that the cluster assignments reported here can be exactly reproduced by any reader running the same code, which is a standard requirement for transparent and scientifically verifiable data analysis (Peng, 2011).

```{r fig12_kmeans}
set.seed(42)
km <- kmeans(scale(pca_data), centers = 3, nstart = 25)

pca_scores          <- as.data.frame(pca$x)
pca_scores$cluster  <- factor(km$cluster)
pca_scores$location <- pca_input$location

label_countries <- c(
  "Germany", "United States", "United Kingdom", "Gibraltar", "Czechia",
  "Brazil", "Israel", "Japan", "Australia", "Turkey",
  "India", "South Africa", "Haiti",
  "Democratic Republic of Congo", "Afghanistan"
)

cluster_colors_named <- c(
  "1" = "#E07000",
  "2" = "#1F3864",
  "3" = "#FFB347"
)

ggplot(pca_scores, aes(PC1, PC2, color = cluster)) +
  stat_ellipse(aes(fill = cluster),
               geom = "polygon", alpha = 0.08,
               level = 0.85, show.legend = FALSE) +
  geom_point(alpha = 0.75, size = 2.8) +
  geom_text_repel(
    data          = pca_scores %>% filter(location %in% label_countries),
    aes(label     = location),
    size          = 2.8, color = "black", fontface = "bold",
    max.overlaps  = 30, box.padding = 0.5, point.padding = 0.3,
    segment.color = "#AAAAAA", segment.size = 0.3,
    min.segment.length = 0.2
  ) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "#CCCCCC", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "#CCCCCC", linewidth = 0.4) +
  scale_color_manual(
    values = cluster_colors_named,
    labels = c(
      "1 -- Upper-middle coverage  (Brazil, Japan, Israel)",
      "2 -- High coverage  (Germany, UK, US)",
      "3 -- Low coverage  (India, DRC, Haiti)"
    )
  ) +
  scale_fill_manual(values = cluster_colors_named) +
  labs(
    title    = "K-means Clusters in PCA Space (k = 3)",
    subtitle = "Countries grouped by vaccination profile -- selected countries labelled",
    x        = "PC1 (76.5% variance -- overall vaccination success)",
    y        = "PC2 (22.8% variance -- daily vaccination intensity)",
    color    = "Cluster",
    caption  = "Ellipses show 85% confidence region per cluster"
  ) +
  theme_orange() +
  theme(
    legend.text     = element_text(size = 8.5),
    legend.key.size = unit(0.9, "lines"),
    plot.caption    = element_text(size = 8, color = "#AAAAAA", hjust = 1)
  )
```

```{r kmeans_summary}
pca_input %>%
  mutate(cluster = factor(km$cluster)) %>%
  group_by(cluster) %>%
  summarise(
    n_countries           = n(),
    mean_fully_vaccinated = mean(people_fully_vaccinated_per_hundred, na.rm = TRUE),
    mean_total            = mean(total_vaccinations_per_hundred, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_fully_vaccinated))
```

K-means clustering identified three distinct country groups. Cluster 2 contains predominantly high-income Western countries (Germany, United Kingdom, United States) with the highest vaccination coverage. Cluster 1 comprises upper-middle income countries (Brazil, Japan, Israel, Australia) with moderate-to-high coverage. Cluster 3 is the largest group and includes predominantly lower-income countries (India, South Africa, Haiti, Democratic Republic of Congo) with low vaccination coverage. The clear separation between clusters confirms that vaccination inequality was structural rather than gradual.

---

## Summary

```{r final_summary}
final_summary <- latest_country %>%
  select(location, date,
         people_fully_vaccinated_per_hundred) %>%
  left_join(
    rollout_speed %>% select(location, reached_50, days_to_50),
    by = "location"
  ) %>%
  left_join(missing_country, by = "location") %>%
  arrange(desc(people_fully_vaccinated_per_hundred))

head(final_summary, 20)
```

COVID-19 vaccination trends differed substantially across countries. High-income countries reached higher coverage earlier, while lower-income countries experienced delayed rollout. The 50% threshold analysis demonstrates that inequality was visible not only in final coverage levels but also in rollout speed. Missing data patterns further suggest that global inequality extends to health data reporting infrastructure -- with consequences for public health planning and resource allocation.

```{r cleanup}
rm(list = ls())
```
