---
title: "vacc_new"
output: html_document
date: "2026-05-09"
---

## Introduction

This project focuses on the analysis of global COVID-19 vaccination data. 
The aim is to explore differences in vaccination rollout speed and vaccination coverage across countries and to evaluate how such data could support public-health monitoring and strategic decision-making.


The first step was to inspect the structure of the dataset. 
The dataset contains daily vaccination records for different countries. 
Basic functions such as `str()`, `summary()`, `head()`, `tail()` and random sampling were used to understand the variables, identify missing values and detect possible formatting problems.

```{r vacc}
View(vaccinations_new)
```


```{r vacc}
head(vaccinations_new)
```

```{r vacc}
tail(vaccinations_new)
```

```{r vacc}
summary(vaccinations_new)
```

```{r vacc}
str(vaccinations_new)
```

```{r vacc}
names(vaccinations_new)
```

```{r vacc}
dim(vaccinations_new)
```

```{r vacc}
colSums(is.na(vaccinations_new))
```

```{r vacc}
vaccinations_new[sample(nrow(vaccinations_new), 1), ]
```

## Data quality issues

During the initial inspection, some variables that should contain numeric values were stored as character variables. In addition, several values contained irregular formats such as `I.32`, `II.55` or `01.XII`. Because the exact meaning of some malformed values was uncertain, these values were interpreted carefully and not aggressively modified.

```{r vacc}
parse_percent_value <- function(x) {

  # pokud je NA nebo prázdné
  if (is.na(x) || x == "") {
    return(NA_real_)
  }

  # odstranění mezer
  x <- str_trim(x)

  # pokud je normální číslo
  if (str_detect(x, "^[0-9]+(\\.[0-9]+)?$")) {
    return(as.numeric(x))
  }

  # oprava levé části
  x <- str_replace(x, "^I\\.", "1.")
  x <- str_replace(x, "^II\\.", "11.")
  x <- str_replace(x, "^III\\.", "111.")
  x <- str_replace(x, "^IV\\.", "14.")
  x <- str_replace(x, "^V\\.", "5.")

  # oprava pravé části
  x <- str_replace(x, "\\.XII$", ".12")
  x <- str_replace(x, "\\.XI$", ".11")
  x <- str_replace(x, "\\.X$", ".10")
  x <- str_replace(x, "\\.IX$", ".9")
  x <- str_replace(x, "\\.VIII$", ".8")
  x <- str_replace(x, "\\.VII$", ".7")
  x <- str_replace(x, "\\.VI$", ".6")
  x <- str_replace(x, "\\.V$", ".5")
  x <- str_replace(x, "\\.IV$", ".4")
  x <- str_replace(x, "\\.III$", ".3")
  x <- str_replace(x, "\\.II$", ".2")
  x <- str_replace(x, "\\.I$", ".1")

  # odstranění nul na začátku
  x <- str_replace(x, "^0+", "")

  # pokud už je validní číslo
  if (str_detect(x, "^[0-9]+(\\.[0-9]+)?$")) {
    return(as.numeric(x))
  }

  # jinak NA
  return(NA_real_)
}

clean_percent_column <- function(x) {
  sapply(x, parse_percent_value)
}
```

```{r vacc}
vaccinations_clean <- vaccinations_new %>%
  mutate(

    date = dmy(date),

    total_vaccinations_per_hundred =
      clean_percent_column(total_vaccinations_per_hundred),

    people_vaccinated_per_hundred =
      clean_percent_column(people_vaccinated_per_hundred),

    people_fully_vaccinated_per_hundred =
      clean_percent_column(people_fully_vaccinated_per_hundred),

    total_boosters_per_hundred =
      clean_percent_column(total_boosters_per_hundred)

  )
```

```{r vacc}
summary(vaccinations_clean)
```

```{r vacc}
vaccinations_clean[sample(nrow(vaccinations_clean), 1), ]

```

# Country-level summary


```{r vacc}
country_summary <- vaccinations_clean %>%
  filter(
    !is.na(people_fully_vaccinated_per_hundred),
    !is.na(daily_vaccinations_per_million)
  ) %>%
  group_by(location) %>%
  summarise(

    max_vaccinated =
      max(people_fully_vaccinated_per_hundred, na.rm = TRUE),

    peak_rate =
      max(daily_vaccinations_per_million, na.rm = TRUE),

    total_vaccinations =
      max(total_vaccinations, na.rm = TRUE)

  )
```

```{r vacc}
country_summary_filtered <- country_summary %>%
  filter(total_vaccinations > 1e6)
```

```{r vacc}
library(ggrepel)
```

```{r vacc}
library(ggplot2)

ggplot(country_summary_filtered,
       aes(x = max_vaccinated,
           y = peak_rate)) +

  geom_point(alpha = 0.6) +

  geom_smooth(method = "lm",
              color = "red") +

  labs(
    title = "Vaccination Performance Across Countries",
    x = "Fully vaccinated per 100 people",
    y = "Peak daily vaccinations per million"
  ) +

  theme_minimal()
```


# The scatter plot shows the relationship between vaccination speed and vaccination coverage. Each point represents one country. The x-axis shows the maximum number of fully vaccinated people per 100 inhabitants, while the y-axis shows the peak daily vaccination rate per million inhabitants. The regression line indicates the general trend between these two variables.

```{r vacc}
library(ggplot2)
library(ggrepel)

# 1) Země pod 100
plot_under_100 <- country_summary_filtered %>%
  filter(max_vaccinated < 100)

ggplot(plot_under_100,
       aes(x = max_vaccinated,
           y = peak_rate)) +

  geom_point(alpha = 0.6) +

  geom_smooth(method = "lm",
              color = "red") +

  geom_text_repel(
    aes(label = location),
    size = 3,
    max.overlaps = Inf
  ) +

  labs(
    title = "Vaccination Performance: Countries Below 100 Fully Vaccinated per 100",
    x = "Fully vaccinated per 100 people",
    y = "Peak daily vaccinations per million"
  ) +

  theme_minimal()
```



# Top high-coverage countries by vaccination speed


```{r vacc}
library(dplyr)
library(ggplot2)

top_high <- country_summary_filtered %>%
  filter(max_vaccinated >= 100) %>%
  arrange(desc(peak_rate)) %>%
  slice(1:20)

ggplot(top_high,
       aes(x = reorder(location, peak_rate),
           y = peak_rate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 20 High-Coverage Countries by Peak Vaccination Rate",
    subtitle = "Countries with at least 70 fully vaccinated people per 100 inhabitants",
    x = "Country",
    y = "Peak daily vaccinations per million"
  ) +
  theme_minimal()
```

```{r vacc}
ggplot(top_high,
       aes(x = reorder(location, peak_rate),
           y = peak_rate)) +
  geom_segment(aes(xend = location, y = 0, yend = peak_rate),
               color = "grey70") +
  geom_point(size = 3, color = "steelblue") +
  coord_flip() +
  labs(
    title = "Top High-Coverage Countries by Vaccination Speed",
    subtitle = "Peak daily vaccinations per million among countries with high coverage",
    x = "Country",
    y = "Peak daily vaccinations per million"
  ) +
  theme_minimal()
```

```{r vacc}

```

```{r vacc}
names(vaccinations_clean)
```

Missing values in vaccination-related variables were expected and primarily reflected delayed vaccination rollout or irregular reporting frequency across countries rather than errors in the dataset itself.

-------------------------------------
12.05.2026
```{r vacc}
vaccinations_clean <- vaccinations_clean %>%
  filter(
    !location %in% c(
      "Europe",
      "European Union",
      "World",
      "Upper middle income",
      "South America",
      "North America",
      "Lower middle income",
      "Low income",
      "High income"
    )
  )
```


```{r vacc}
missing_summary <- data.frame(
  variable = names(vaccinations_clean),
  missing_count = colSums(is.na(vaccinations_clean)),
  missing_percent = round(colMeans(is.na(vaccinations_clean)) * 100, 2)
)

missing_summary
```


The analysis of missing values revealed substantial differences in reporting completeness across variables. Booster-related variables contained the highest proportion of missing observations, likely because booster vaccination campaigns started later and were not consistently reported by all countries.

Variables related to daily vaccination activity contained relatively few missing values, making them more reliable for comparative analysis.

The observed missing values are interpreted primarily as a consequence of delayed vaccination rollout and inconsistent reporting frequency rather than errors in the dataset itself.
```{r vacc}
ggplot(missing_summary,
       aes(x = reorder(variable, missing_percent),
           y = missing_percent)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Missing Values by Variable",
    x = "Variable",
    y = "Missing values (%)"
  ) +
  theme_minimal()
```

```{r vacc}
vaccination_start <- vaccinations_clean %>%
  filter(!is.na(total_vaccinations),
         total_vaccinations > 0) %>%
  group_by(location) %>%
  summarise(
    first_vaccination_date = min(date, na.rm = TRUE)
  ) %>%
  arrange(first_vaccination_date)

vaccination_start
```

```{r vacc}
vaccination_start %>%
  slice(1:20) %>%
  ggplot(aes(x = reorder(location, first_vaccination_date),
             y = first_vaccination_date)) +
  geom_point(size = 3, color = "steelblue") +
  coord_flip() +
  labs(
    title = "First Reported Vaccination Date by Country",
    x = "Country",
    y = "First vaccination date"
  ) +
  theme_minimal()
```


```{r vacc}
country_summary <- vaccinations_clean %>%
  group_by(location) %>%
  summarise(
    max_total_vaccinations_per_100 =
      max(total_vaccinations_per_hundred, na.rm = TRUE),
    
    max_fully_vaccinated_per_100 =
      max(people_fully_vaccinated_per_hundred, na.rm = TRUE),
    
    peak_daily_vaccinations_per_million =
      max(daily_vaccinations_per_million, na.rm = TRUE),
    
    total_vaccinations =
      max(total_vaccinations, na.rm = TRUE)
  ) %>%
  filter(
    is.finite(max_total_vaccinations_per_100),
    is.finite(max_fully_vaccinated_per_100),
    is.finite(peak_daily_vaccinations_per_million)
  )
```

```{r vacc}
country_summary_filtered <- country_summary %>%
  filter(total_vaccinations > 1000000)

ggplot(country_summary_filtered,
       aes(x = max_fully_vaccinated_per_100,
           y = peak_daily_vaccinations_per_million)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Vaccination Speed and Final Vaccination Coverage",
    x = "Fully vaccinated per 100 people",
    y = "Peak daily vaccinations per million"
  ) +
  theme_minimal()
```

```{r vacc}
top_speed <- country_summary_filtered %>%
  arrange(desc(peak_daily_vaccinations_per_million)) %>%
  slice(1:20)

ggplot(top_speed,
       aes(x = reorder(location, peak_daily_vaccinations_per_million),
           y = peak_daily_vaccinations_per_million)) +
  geom_segment(aes(xend = location,
                   y = 0,
                   yend = peak_daily_vaccinations_per_million),
               color = "grey70") +
  geom_point(size = 3, color = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 20 Countries by Peak Vaccination Speed",
    x = "Country",
    y = "Peak daily vaccinations per million"
  ) +
  theme_minimal()
```


```{r vacc}
cor(
  country_summary_filtered$max_fully_vaccinated_per_100,
  country_summary_filtered$peak_daily_vaccinations_per_million,
  use = "complete.obs"
)
```

A linear regression model was used to examine the relationship between vaccination rollout speed and final vaccination coverage across countries.

The results indicate a statistically significant positive relationship between the peak daily vaccination rate and the proportion of fully vaccinated individuals per 100 inhabitants (p < 0.001).

However, the model explained approximately 17% of the variability in vaccination coverage (R² = 0.174), suggesting that additional factors beyond rollout speed also influence vaccination outcomes across countries.

```{r vacc}
model <- lm(
  max_fully_vaccinated_per_100 ~ peak_daily_vaccinations_per_million,
  data = country_summary_filtered
)

summary(model)
```
#time analysis
The time-series analysis reveals substantial differences in vaccination rollout strategies across countries.

Israel exhibited the fastest early vaccination rollout and achieved high vaccination coverage within a relatively short period of time. The United States also demonstrated rapid vaccination progress, although with a more gradual increase compared to Israel.

Germany showed a slower initial rollout followed by a later acceleration phase. In contrast, India experienced a considerably slower vaccination campaign and achieved substantially lower vaccination coverage during the observed period.

These differences highlight the impact of country-specific factors such as population size, healthcare infrastructure, logistics, and vaccine availability.
```{r vacc}
Sys.setlocale("LC_TIME", "English")

selected_countries <- vaccinations_clean %>%
  filter(location %in% c("United States",
                         "Germany",
                         "India",
                         "Israel"))

ggplot(selected_countries,
       aes(x = date,
           y = people_fully_vaccinated_per_hundred,
           color = location)) +

  geom_line(size = 1) +

  labs(
    title = "Vaccination Progress Over Time",
    x = "Date",
    y = "Fully vaccinated per 100 people",
    color = "Country"
  ) +

  theme_minimal()
```


```{r vacc}

```

```{r vacc}

```

```{r vacc}

```


```{r vacc}

```

```{r vacc}

```

```{r vacc}

```


```{r vacc}

```

```{r vacc}

```

```{r vacc}

```


```{r vacc}

```

```{r vacc}

```

```{r vacc}

```


