
```{r}
library(dplyr)
library(stringr)
library(lubridate)
```

```{r}
clean_percent_column <- function(x) {

  x %>%
    str_trim() %>%
    
    # fix OCR-like patterns
    str_replace_all("^IV\\.", "14.") %>%
    str_replace_all("^III\\.", "111.") %>%
    str_replace_all("^II\\.", "11.") %>%
    str_replace_all("^I\\.", "1.") %>%
    
    # fix suffix cases
    str_replace_all("\\.VII$", ".7") %>%
    str_replace_all("\\.VI$", ".6") %>%
    str_replace_all("\\.V$", ".5") %>%
    str_replace_all("\\.IV$", ".4") %>%
    str_replace_all("\\.III$", ".3") %>%
    str_replace_all("\\.II$", ".2") %>%
    str_replace_all("\\.I$", ".1") %>%
    
    # remove leading zeros like 04.7 → 4.7
    str_replace("^0+", "") %>%
    
    na_if("") %>%
    as.numeric()
}
```

```{r}
vaccinations_clean <- vaccinations %>%
  mutate(
    date = dmy(date),
    total_vaccinations_per_hundred = clean_percent_column(total_vaccinations_per_hundred),
    people_vaccinated_per_hundred = clean_percent_column(people_vaccinated_per_hundred),
    people_fully_vaccinated_per_hundred = clean_percent_column(people_fully_vaccinated_per_hundred),
    total_boosters_per_hundred = clean_percent_column(total_boosters_per_hundred)
  )
```
```{r vacc}
vaccinations_clean <- vaccinations_clean %>%
  filter(
    !location %in% c(
      "Europe",
      "European Union",
      "World",
      "Upper middle income",
      "South America",
      "North America",
      "Lower middle income",
      "Low income",
      "High income",
      "Lower middle income",      
      "Asia",
      "Africa",
      "Low income",
      "South America",
      "Upper middle income",
      "High income",
      "South Africa",
      "European Union",
      "Europe",
      "North America",
      "Oceania",
      "World",
      "Lower middle income",
      "United Kingdom"
      
    )
  )
```



```{r}
vaccinations %>%
  mutate(cleaned = clean_percent_column(total_vaccinations_per_hundred)) %>%
  filter(
    !is.na(total_vaccinations_per_hundred) &
    is.na(cleaned)
  ) %>%
  distinct(total_vaccinations_per_hundred)
```


```{r}
library(dplyr)
library(stringr)
library(lubridate)

roman_to_number <- function(x) {
  case_when(
    x == "I" ~ 1,
    x == "II" ~ 2,
    x == "III" ~ 3,
    x == "IV" ~ 4,
    x == "V" ~ 5,
    x == "VI" ~ 6,
    x == "VII" ~ 7,
    x == "VIII" ~ 8,
    x == "IX" ~ 9,
    x == "X" ~ 10,
    x == "XI" ~ 11,
    x == "XII" ~ 12,
    TRUE ~ NA_real_
  )
}

parse_mixed_number <- function(x) {
  if (is.na(x) || x == "") return(NA_real_)
  
  x <- str_trim(x)
  
  # normal numeric value
  if (str_detect(x, "^[0-9]+(\\.[0-9]+)?$")) {
    return(as.numeric(x))
  }
  
  # values like I.51, II.55, IV.22, V.14, 01.VIII, 03.XI
  if (str_detect(x, "^[A-Z0-9]+\\.[A-Z0-9]+$")) {
    parts <- str_split(x, "\\.", simplify = TRUE)
    left <- parts[1]
    right <- parts[2]
    
    left_num <- if (str_detect(left, "^[0-9]+$")) {
      as.numeric(left)
    } else {
      roman_to_number(left)
    }
    
    right_num <- if (str_detect(right, "^[0-9]+$")) {
      as.numeric(right)
    } else {
      roman_to_number(right)
    }
    
    if (is.na(left_num) || is.na(right_num)) {
      return(NA_real_)
    }
    
    # decimal reconstruction: V.14 -> 5.14, 01.VIII -> 1.08
    return(left_num + right_num / 100)
  }
  
  NA_real_
}

clean_percent_column <- function(x) {
  sapply(x, parse_mixed_number)
}
```


```{r}

```


```{r}
vaccinations_clean <- vaccinations %>%
  mutate(
    date = dmy(date),
    total_vaccinations_per_hundred = clean_percent_column(total_vaccinations_per_hundred),
    people_vaccinated_per_hundred = clean_percent_column(people_vaccinated_per_hundred),
    people_fully_vaccinated_per_hundred = clean_percent_column(people_fully_vaccinated_per_hundred),
    total_boosters_per_hundred = clean_percent_column(total_boosters_per_hundred)
  )
```


```{r}
problem_values <- vaccinations %>%
  mutate(cleaned = clean_percent_column(total_vaccinations_per_hundred)) %>%
  filter(!is.na(total_vaccinations_per_hundred) & is.na(cleaned)) %>%
  distinct(total_vaccinations_per_hundred)

problem_values
```


```{r}
summary(vaccinations_clean$total_vaccinations_per_hundred)
```
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   0.00    3.96   20.22   37.18   60.86  233.93   18225  
= number of doses administered per 100 population.

The variable total_vaccinations_per_hundred measures the number of vaccine doses administered per 100 inhabitants. 
The mean value (37.18) exceeds the median (20.22), indicating a right-skewed distribution. 
This suggests that while most observations have relatively low to moderate vaccination coverage, a smaller number of countries achieved substantially higher vaccination rates. 
The highest recorded value was 233.93 doses per 100 inhabitants, largely due to the administration of booster doses.
```{r}
hist(
  vaccinations_clean$total_vaccinations_per_hundred,
  breaks = 30,
  main = "Distribution of Total Vaccinations per Hundred",
  xlab = "Vaccinations per Hundred"
)
```
The histogram reveals a highly positively skewed distribution of total_vaccinations_per_hundred. 
Most observations are clustered at low vaccination levels, whereas a smaller number of countries achieved substantially higher vaccination coverage. This pattern is reflected by the mean (37.18) exceeding the median (20.22). 
The maximum value of 233.93 doses per 100 inhabitants suggests extensive booster vaccination campaigns in some countries. 
Additionally, 18,225 observations were missing for this variable.



```{r}
summary(vaccinations_clean$people_fully_vaccinated_per_hundred)
```
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   0.00    1.57    7.77   16.21   25.27  116.40   22281 

The variable people_fully_vaccinated_per_hundred exhibits a strongly positively skewed distribution. 
The mean (16.21) exceeds the median (7.77), suggesting that a relatively small number of observations with high vaccination coverage increase the overall average. 
Most observations are concentrated at lower values, indicating that many countries were still in the early phases of their vaccination campaigns during a substantial part of the observed period. 
The highest recorded value was 116.40 fully vaccinated individuals per 100 inhabitants. 
In addition, 22,281 observations contained missing values.


```{r}
hist(
  vaccinations_clean$people_fully_vaccinated_per_hundred,
  breaks = 30,
  main = "Distribution of People fully vaccinated per Hundred",
  xlab = "Vaccinations per Hundred"
)
```
The histogram of people_fully_vaccinated_per_hundred reveals a strongly right-skewed distribution. 
Most observations are concentrated at very low vaccination levels, with the highest frequencies occurring below 10 fully vaccinated individuals per 100 inhabitants. 
The distribution exhibits a long right tail, indicating the presence of a smaller number of observations with substantially higher vaccination coverage. 
This pattern is reflected in the summary statistics, where the mean (16.21) exceeds the median (7.77). 
The maximum recorded value was 116.40 fully vaccinated individuals per 100 inhabitants, which may be attributed to population estimation differences and reporting adjustments. 
Furthermore, 22,281 observations contained missing values.
   
```{r}
Sys.getlocale("LC_TIME")
```
   
   
```{r}
library(dplyr)
library(ggplot2)

vaccinations_clean %>%
  group_by(date) %>%
  summarise(
    avg_fully_vaccinated =
      mean(people_fully_vaccinated_per_hundred,
           na.rm = TRUE)
  ) %>%
  ggplot(aes(x = date,
             y = avg_fully_vaccinated)) +
  geom_line(color = "steelblue",
            linewidth = 1) +
  scale_x_date(date_labels = "%b %Y") +
  labs(
    title = "Average Full Vaccination Coverage Over Time",
    x = "Date",
    y = "Fully Vaccinated per 100"
  ) +
  theme_minimal()
```
The average share of fully vaccinated individuals increased steadily throughout the observed period. 
Initially, vaccination coverage remained close to zero, reflecting the early phase of vaccine deployment. 
As vaccination campaigns expanded across countries, the average proportion of fully vaccinated individuals rose rapidly, reaching approximately 40 fully vaccinated persons per 100 inhabitants by the end of the observation period. 
Minor fluctuations in the trend are likely caused by differences in reporting schedules and data availability across countries. 
Overall, the figure demonstrates the global progress of COVID-19 vaccination campaigns over time.


```{r}
selected <- c(
  "United States",
  "Germany",
  "France",
  "Czechia",
  "India"
)

vaccinations_clean %>%
  filter(location %in% selected) %>%
  ggplot(aes(
    x = date,
    y = people_fully_vaccinated_per_hundred,
    color = location
  )) +
  geom_line() +
  scale_x_date(date_labels = "%b %Y")+
  theme_minimal() +
  labs(
    title = "Vaccination Progress in Selected Countries",
    x = "Date",
    y = "Fully Vaccinated per 100"
  )
```

```{r}
top10 <- vaccinations_clean %>%
  group_by(location) %>%
  summarise(
    max_vacc = max(
      people_fully_vaccinated_per_hundred,
      na.rm = TRUE
    )
  ) %>%
  arrange(desc(max_vacc)) %>%
  slice(1:10) %>%
  pull(location)

vaccinations_clean %>%
  filter(location %in% top10) %>%
  ggplot(aes(
    x = date,
    y = people_fully_vaccinated_per_hundred,
    color = location
  )) +
  geom_line() +
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y")+
  labs(
    title = "Top 10 Countries by Full Vaccination Coverage",
    x = "Date",
    y = "Fully Vaccinated per 100"
  )
```


```{r}

```

```{r}
top_countries <- vaccinations_clean %>%
  group_by(location) %>%
  summarise(max_vac = max(total_vaccinations_per_hundred, na.rm = TRUE)) %>%
  arrange(desc(max_vac)) %>%
  slice(1:5) %>%
  pull(location)
```

```{r}
vaccinations_clean %>%
  filter(location %in% top_countries,
         !is.na(total_vaccinations_per_hundred)) %>%
  ggplot(aes(x = date, y = total_vaccinations_per_hundred, color = location)) +
  geom_line(size = 0.7) +
  labs(
    title = "Top performers",
    x = "Date",
    y = "Total vaccinations per 100 people"
  ) +
  theme_minimal()
```

```{r}
theme_minimal() +
theme(
  plot.title = element_text(size = 14, face = "bold"),
  legend.position = "right"
)
```

```{r}
country_summary <- vaccinations_clean %>%
  filter(
    !is.na(people_fully_vaccinated_per_hundred),
    !is.na(daily_vaccinations_per_million)
  ) %>%
  group_by(location) %>%
  summarise(
    max_vaccinated = max(people_fully_vaccinated_per_hundred, na.rm = TRUE),
    avg_daily_vaccinations = mean(daily_vaccinations_per_million, na.rm = TRUE)
  )
```

```{r}

corr <- cor(
  country_summary$max_vaccinated,
  country_summary$avg_daily_vaccinations,
  use = "complete.obs"
)


ggplot(country_summary,
       aes(x = max_vaccinated,
           y = avg_daily_vaccinations)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  annotate(
    "text",
    x = 80,
    y = max(country_summary$avg_daily_vaccinations, na.rm = TRUE),
    label = paste("r =", round(corr, 3))
  ) +
  labs(
    title = "Vaccination Coverage vs Daily Vaccination Rate",
    x = "Fully vaccinated per 100 people",
    y = "Daily vaccinations per million people"
  ) +
  theme_minimal()
```
There is a moderate positive correlation between the average daily vaccination rate and the maximum vaccination coverage achieved by a country (r = 0.517).

```{r}
cor.test(
  country_summary$max_vaccinated,
  country_summary$avg_daily_vaccinations
)
```

A scatter plot was used to examine the relationship between the average daily vaccination rate and the maximum vaccination coverage achieved by each country. 
The analysis revealed a moderate positive correlation (r = 0.517, p < 0.001), indicating that countries with higher daily vaccination rates generally tended to achieve higher vaccination coverage. 
The fitted regression line confirms this positive trend. 
However, the substantial dispersion of observations suggests that vaccination speed alone does not fully explain the differences in vaccination coverage across countries. 
Additional factors, such as vaccine availability, public acceptance, and national vaccination policies, likely contributed to the observed variation.






```{r}
install.packages("ggrepel")

```

```{r}
library(ggrepel)
label_data <- country_summary %>%
  arrange(desc(avg_daily_vaccinations)) %>%
  slice(1:6)  
```

```{r}
ggplot(country_summary,
       aes(x = max_vaccinated, y = avg_daily_vaccinations)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  
  geom_text_repel(
    data = label_data,
    aes(label = location),
    size = 4
  ) +
  
  labs(
    title = "Vaccination Coverage vs Daily Vaccination Rate",
    x = "Fully vaccinated per 100 people",
    y = "Daily vaccinations per million people"
  ) +
  
  theme_minimal()
```
The scatter plot shows a moderate positive relationship between maximum vaccination coverage and the average daily vaccination rate. 
Several small territories, such as Niue, Bhutan, the Falkland Islands, Nauru, Pitcairn, and the Cook Islands, appear as clear outliers with unusually high daily vaccination rates per million people. 
This is likely influenced by their small population sizes, where even a relatively small absolute number of vaccinations can result in very high per-million values. 
Despite these outliers, the regression line indicates a positive overall trend: countries with higher daily vaccination rates generally tended to achieve higher vaccination coverage.



```{r}
country_summary <- vaccinations_clean %>%
  filter(
    !is.na(people_fully_vaccinated_per_hundred),
    !is.na(daily_vaccinations_per_million)
  ) %>%
  group_by(location) %>%
  summarise(
    max_vaccinated = max(people_fully_vaccinated_per_hundred, na.rm = TRUE),
    peak_rate = max(daily_vaccinations_per_million, na.rm = TRUE),
    total_vaccinations = max(total_vaccinations, na.rm = TRUE)
  ) %>%
  filter(total_vaccinations > 1e6)   
```

```{r}
library(ggplot2)

ggplot(country_summary,
       aes(x = max_vaccinated, y = peak_rate)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", color = "red") +
  
  labs(
    title = "Vaccination Performance Across Countries (Filtered)",
    x = "Fully vaccinated per 100 people",
    y = "Peak daily vaccinations per million"
  ) +
  
  theme_minimal()
```
After filtering out countries and territories with fewer than one million total vaccinations, the relationship between vaccination coverage and peak daily vaccination rate remains positive. 
The regression line indicates that countries with higher peak vaccination capacity generally tended to achieve higher full vaccination coverage. 
However, the dispersion of points shows that peak vaccination speed alone does not fully explain final vaccination coverage. 
Other factors, such as vaccine availability, public trust, policy decisions, and healthcare infrastructure, likely influenced the observed differences between countries.


```{r}
library(ggrepel)

label_data <- country_summary %>%
  arrange(desc(peak_rate)) %>%
  slice(1:8)

ggplot(country_summary,
       aes(x = max_vaccinated, y = peak_rate)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  
  geom_text_repel(
    data = label_data,
    aes(label = location),
    size = 4
  ) +
  
  theme_minimal()
```
The scatter plot compares the maximum vaccination coverage achieved by each country with its peak daily vaccination rate. 
A positive relationship can be observed, suggesting that countries capable of administering vaccines at a higher daily rate generally achieved higher vaccination coverage. 
Several countries, including Bhutan, Mongolia, Sri Lanka, Cuba, and Israel, stand out as notable outliers due to exceptionally high vaccination rates during the peak of their vaccination campaigns. 
These observations indicate that rapid vaccine deployment was associated with successful vaccination outcomes, although considerable variation between countries remains evident. 
Such variation suggests that additional factors, including vaccine availability, public acceptance, and government policies, also played an important role in determining final vaccination coverage.



```{r}
"
corelace
boxplots
how many na
try to find which columns have NA
-find relation about NA
na are ussualy more in older people

remove europe, european!

understand why there are NA
plot it with NA or without NA
understand the basic coefitients, what does the baic values means, what does it indicate

if are there any exceptions in data set
plotting 
does pca and clustering be useful for like (country similialirity, country similarity, welth of country)

when making model, understand the coeffitiens, how to read the residuals
"

```




```{r}

colSums(is.na(vaccinations_clean))
```
```{r}
na_counts <- colSums(is.na(vaccinations_clean))

round(
  na_counts / nrow(vaccinations_clean) * 100,
  2
)
```
Missing value analysis revealed substantial differences in data completeness across variables. 
The identifier variables (location, iso_code, and date) contained no missing values. 
Daily vaccination indicators (daily_vaccinations and daily_vaccinations_per_million) were nearly complete, with only 0.61% missing observations. 
In contrast, cumulative vaccination measures such as total_vaccinations, people_vaccinated, and people_fully_vaccinated contained between 42% and 51% missing values. 
These missing values are likely associated with countries that had not yet started vaccination campaigns or did not consistently report cumulative vaccination statistics. 
The highest proportion of missing values was observed in booster-related variables (total_boosters and total_boosters_per_hundred), where 98.82% of observations were missing. 
This is expected because booster vaccination programs were introduced later during the pandemic and were not reported consistently across all countries. 
Therefore, the observed missing values primarily reflect differences in reporting practices and vaccination rollout timelines rather than data quality problems.

```{r}
vaccinations_clean %>%
  mutate(
    missing = is.na(people_fully_vaccinated_per_hundred)
  ) %>%
  group_by(date) %>%
  summarise(
    missing_pct = mean(missing) * 100
  )
```


```{r}
ggplot(
  vaccinations_clean,
  aes(y = people_fully_vaccinated_per_hundred)
) +
  geom_boxplot()
```
The boxplot confirms that the distribution of people_fully_vaccinated_per_hundred is highly right-skewed. 
The median vaccination coverage was approximately 7.77 fully vaccinated individuals per 100 inhabitants, while the interquartile range extended from 1.57 to 25.27. 
A large number of observations were identified as statistical outliers above approximately 60 fully vaccinated individuals per 100 inhabitants. 
These outliers likely correspond to countries that achieved exceptionally high vaccination coverage during the observed period. 
The presence of numerous upper-tail outliers is consistent with the right-skewed distribution observed in the histogram and summary statistics.



```{r}
outliers <- boxplot.stats(
  vaccinations_clean$people_fully_vaccinated_per_hundred
)$out

vaccinations_clean %>%
  filter(
    people_fully_vaccinated_per_hundred %in% outliers
  ) %>%
  select(location,
         people_fully_vaccinated_per_hundred) %>%
  distinct() %>%
  arrange(desc(people_fully_vaccinated_per_hundred))
```



```{r}
country_summary %>%
  arrange(desc(peak_rate)) %>%
  head(10)
```
Outlier analysis revealed several countries with exceptionally high peak vaccination rates. 
Bhutan exhibited the highest peak rate, reaching 78,604 daily vaccinations per million inhabitants, followed by Mongolia (37,684) and Sri Lanka (22,125). 
These values are substantially higher than those observed in most countries and can largely be explained by intensive vaccination campaigns combined with relatively small population sizes. 
Since these observations correspond to real vaccination efforts rather than measurement errors, they were retained in the dataset and interpreted as meaningful extreme cases rather than anomalies requiring removal.



```{r}
as.data.frame(
  vaccinations_clean %>%
    group_by(location) %>%
    summarise(
      na_pct = mean(
        is.na(people_fully_vaccinated_per_hundred)
      ) * 100
    ) %>%
    arrange(desc(na_pct))
)


na_country <- vaccinations_clean %>%
  group_by(location) %>%
  summarise(
    na_pct = mean(
      is.na(people_fully_vaccinated_per_hundred)
    ) * 100
  ) %>%
  arrange(desc(na_pct))

View(na_country)
```

```{r}
na_country <- vaccinations_clean %>%
  group_by(location) %>%
  summarise(
    na_pct = mean(
      is.na(people_fully_vaccinated_per_hundred)
    ) * 100
  ) %>%
  arrange(desc(na_pct))

ggplot(
  na_country %>% slice(1:20),
  aes(
    x = reorder(location, na_pct),
    y = na_pct
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Countries with Highest Percentage of Missing Values",
    x = "Country",
    y = "Missing Values (%)"
  ) +
  theme_minimal()
```
```{r}
na_country %>%
  arrange(na_pct) %>%
  head(20)
```



```{r}
summary(na_country$na_pct)
```
For at least half of the countries, more than 62% of observations for people_fully_vaccinated_per_hundred are missing.


```{r}
ggplot(
  na_country,
  aes(x = na_pct)
) +
  geom_histogram(
    bins = 30,
    fill = "steelblue"
  ) +
  labs(
    title = "Distribution of Missing Values Across Countries",
    x = "Percentage of Missing Values",
    y = "Number of Countries"
  ) +
  theme_minimal()
```
The distribution of missing values across countries is highly uneven. 
While some countries reported vaccination statistics almost completely, others contained very high proportions of missing observations. 
The median percentage of missing values was 62.77%, indicating that more than half of the countries had missing data for the majority of observations. 
The histogram suggests the presence of two distinct groups: countries with relatively complete reporting and countries with very limited reporting. 
This pattern indicates that missing values are not randomly distributed but are strongly associated with country-specific reporting practices and data availability.


```{r}
vaccinations_clean %>%
  filter(location == "Cote d'Ivoire") %>%
  summarise(
    total = n(),
    na_count = sum(is.na(people_fully_vaccinated_per_hundred))
  )
```


```{r}
country_features <- vaccinations_clean %>%
  group_by(location) %>%
  summarise(
    max_vaccinated =
      ifelse(
        all(is.na(people_fully_vaccinated_per_hundred)),
        NA,
        max(
          people_fully_vaccinated_per_hundred,
          na.rm = TRUE
        )
      )
  )
```

```{r}
country_features <- vaccinations_clean %>%
  group_by(location) %>%
  summarise(
    max_vaccinated =
      if (all(is.na(people_fully_vaccinated_per_hundred))) {
        NA_real_
      } else {
        max(people_fully_vaccinated_per_hundred, na.rm = TRUE)
      },

    peak_rate =
      if (all(is.na(daily_vaccinations_per_million))) {
        NA_real_
      } else {
        max(daily_vaccinations_per_million, na.rm = TRUE)
      },

    total_vaccinations =
      if (all(is.na(total_vaccinations))) {
        NA_real_
      } else {
        max(total_vaccinations, na.rm = TRUE)
      },

    .groups = "drop"
  ) %>%
  filter(
    !is.na(max_vaccinated),
    !is.na(peak_rate),
    !is.na(total_vaccinations)
  )
```


```{r}
vaccinations_clean %>%
  group_by(location) %>%
  summarise(
    all_na_peak_rate = all(is.na(daily_vaccinations_per_million)),
    .groups = "drop"
  ) %>%
  filter(all_na_peak_rate)
```

```{r}
country_features <- vaccinations_clean %>%
  group_by(location) %>%
  summarise(
    max_vaccinated =
      if (all(is.na(people_fully_vaccinated_per_hundred))) {
        NA_real_
      } else {
        max(people_fully_vaccinated_per_hundred, na.rm = TRUE)
      },

    peak_rate =
      if (all(is.na(daily_vaccinations_per_million))) {
        NA_real_
      } else {
        max(daily_vaccinations_per_million, na.rm = TRUE)
      },

    total_vaccinations =
      if (all(is.na(total_vaccinations))) {
        NA_real_
      } else {
        max(total_vaccinations, na.rm = TRUE)
      },

    .groups = "drop"
  ) %>%
  filter(
    !is.na(max_vaccinated),
    !is.na(peak_rate),
    !is.na(total_vaccinations)
  )
```


```{r}
View(country_features)
summary(country_features)
nrow(country_features)
```
```{r}
bad_countries <- c(
  "Cote d'Ivoire",
  "Ethiopia",
  "Kiribati",
  "Madagascar",
  "Tokelau",
  "Tuvalu",
  "Bonaire Sint Eustatius and Saba",
  "Turkmenistan"
)
```


```{r}
country_features <- vaccinations_clean %>%
  filter(!location %in% bad_countries) %>%
  group_by(location) %>%
  summarise(
    max_fully_vaccinated =
      max(people_fully_vaccinated_per_hundred, na.rm = TRUE),

    max_people_vaccinated =
      max(people_vaccinated_per_hundred, na.rm = TRUE),

    max_total_vaccinations =
      max(total_vaccinations_per_hundred, na.rm = TRUE),

    peak_daily_rate =
      max(daily_vaccinations_per_million, na.rm = TRUE),

    mean_daily_rate =
      mean(daily_vaccinations_per_million, na.rm = TRUE),

    .groups = "drop"
  )
```



```{r}
country_features <- country_features %>%
  filter(if_all(-location, is.finite))
```



```{r}
pca_data <- country_features %>%
  select(
    max_fully_vaccinated,
    max_people_vaccinated,
    max_total_vaccinations,
    peak_daily_rate,
    mean_daily_rate
  )
```

```{r}
pca <- prcomp(
  pca_data,
  center = TRUE,
  scale. = TRUE
)

summary(pca)
```



```{r}
pca$rotation
```
The PCA results showed that the first principal component (PC1) explained 80.27% of the total variance in the dataset, while the second principal component (PC2) explained an additional 13.98%.
Together, the first two components accounted for 94.25% of the total variability.

This indicates that most differences between countries can be effectively represented in a two-dimensional PCA space.

```{r}
scores <- as.data.frame(pca$x)
scores$location <- country_features$location

ggplot(scores,
       aes(PC1, PC2)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "PCA of Vaccination Characteristics"
  )
```
The PCA scatter plot reveals a clear gradient along PC1, reflecting differences in overall vaccination performance. 
Most countries form a dense central cluster, while several countries appear as outliers on the right side of the plot. 
These outliers correspond to countries with exceptionally successful vaccination campaigns and/or unusually high vaccination rates.

The plot suggests that vaccination performance varies continuously across countries rather than forming sharply separated groups. 
However, a small number of countries exhibit substantially different vaccination characteristics compared to the majority of observations.

```{r}

```



```{r}
scores %>%
  arrange(desc(PC1)) %>%
  select(location, PC1, PC2) %>%
  head(10)
```

```{r}
scores %>%
  arrange(desc(PC2)) %>%
  select(location, PC1, PC2) %>%
  head(10)
```
Countries with the highest PC1 scores included Bhutan, Gibraltar, Niue, Falkland Islands, Pitcairn, and the United Arab Emirates. 
Since PC1 represents overall vaccination performance, these countries achieved exceptionally high values across multiple vaccination indicators. 
Many of these observations correspond to small island territories, where vaccination metrics expressed per hundred inhabitants or per million inhabitants can reach unusually high values.

Countries with the highest PC2 scores included Scotland, Wales, England, Portugal, Malta, and Qatar. 
These countries achieved high vaccination coverage while not necessarily exhibiting the most extreme vaccination rates. 
This suggests that PC2 captures differences between vaccination speed and final vaccination coverage.

```{r}
scaled_data <- scale(pca_data)

km <- kmeans(
  scaled_data,
  centers = 3,
  nstart = 50
)
```


```{r}
ggplot(scores, aes(PC1, PC2, color = cluster)) +
  geom_point(size = 2) +
  theme_minimal()
```
K-means clustering was performed on the standardized vaccination indicators. 
PCA was subsequently used to visualize the resulting clusters in a two-dimensional space.

```{r}
country_features$cluster <- factor(km$cluster)

country_features %>%
  group_by(cluster) %>%
  summarise(
    avg_fully_vaccinated =
      mean(max_fully_vaccinated),

    avg_peak_rate =
      mean(peak_daily_rate),

    avg_total_vaccinations =
      mean(max_total_vaccinations)
  )
```
```{r}
km <- kmeans(
  pca$x[,1:2],
  centers = 3
)
```

```{r}
table(km$cluster)
```

```{r}

```

