---
title: "ARIMA Forecasting: Temporal Inequality in Vaccination Progress"
author: "Eva Štětinová & Ayushman Mandhotra"
date: "`r Sys.Date()`"
output: html_document
---

<!--
STANDALONE NOTEBOOK. This file used to depend on objects created only inside
final_complete.Rmd (vaccinations_country) and referenced an undefined colour
(red_base). Both are fixed here: this file loads and cleans the raw CSV
itself, so it can be knit on its own, and defines its own full colour
palette. It should be run from the same working directory as
vaccination_pipeline.Rmd and vaccinations.csv.
-->

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 7.5, fig.height = 4.5,
  fig.path = "figures/arima/", dev = "png", dpi = 300
)

library(dplyr)
library(tibble)
library(ggplot2)
library(forecast)
library(tseries)
library(zoo)

# ---- Colour palette (kept identical to vaccination_pipeline.Rmd for visual
# consistency across the two deliverables) ---------------------------------
teal_dark   <- "#015c6d"
teal_mid    <- "#017a8f"
teal_base   <- "#019ab5"
teal_light  <- "#4dc2d6"
teal_pale   <- "#a8e2ec"
teal_ghost  <- "#d9f2f6"

red_dark    <- "#c02200"
red_mid     <- "#fd300b"
red_base    <- "#fd5a38"
red_light   <- "#fe8a70"

theme_teal <- function() {
  theme_minimal() +
    theme(
      plot.title       = element_text(face = "bold", colour = "black"),
      plot.subtitle    = element_text(colour = "black"),
      axis.title       = element_text(colour = teal_dark),
      axis.text        = element_text(colour = "#555555"),
      legend.title     = element_text(colour = teal_dark, face = "bold"),
      panel.grid.major = element_line(colour = teal_ghost),
      panel.grid.minor = element_blank()
    )
}
```

---

## Data Loading and Cleaning

This mirrors the loading step in `vaccination_pipeline.Rmd`. It is repeated
here (rather than assumed) so that this notebook can be knit independently.

```{r load_data}
vaccinations <- read.csv("vaccinations.csv", sep = ",", stringsAsFactors = FALSE)

vaccinations_clean <- vaccinations %>%
  mutate(
    date = as.Date(date),
    across(
      c(total_vaccinations_per_hundred,
        people_vaccinated_per_hundred,
        people_fully_vaccinated_per_hundred,
        total_boosters_per_hundred),
      ~ suppressWarnings(as.numeric(.x))
    )
  )

regional_aggregates <- c(
  "World", "Europe", "European Union", "Asia", "Africa",
  "North America", "South America", "Oceania",
  "High income", "Upper middle income",
  "Lower middle income", "Low income"
)

vaccinations_country <- vaccinations_clean %>%
  filter(!location %in% regional_aggregates)
```

## Preparing Clean, Gap-Free Time Series

The raw reporting calendar has gaps: not every country reports on every
calendar day. For Germany, every date in range has a value, but for India
and Bangladesh there are real gaps (multi-day jumps between reported
observations). Passing a vector with silent gaps straight into `ts(...,
frequency = 7)` is a bug — it assumes row *position*, not the actual
calendar date, defines the weekly cycle, which misaligns the seasonal index
whenever a country has skipped days. This function builds a complete daily
calendar for a country and linearly interpolates missing values (with
`rule = 2` so the first/last known value is carried to the edges) before
any time series object is created.

```{r gap_fill_function}
prepare_country_series <- function(data, loc, value_col) {
  d <- data %>%
    filter(location == loc, !is.na(.data[[value_col]])) %>%
    arrange(date) %>%
    select(date, value = all_of(value_col))

  full_dates <- tibble(date = seq(min(d$date), max(d$date), by = "day"))

  d_full <- full_dates %>% left_join(d, by = "date")

  value_zoo <- zoo::zoo(d_full$value, order.by = d_full$date)
  d_full$value <- as.numeric(zoo::na.approx(value_zoo, rule = 2))

  d_full
}

n_gaps <- function(data, loc, value_col) {
  d <- data %>% filter(location == loc, !is.na(.data[[value_col]])) %>% arrange(date)
  sum(diff(d$date) > 1)
}

for (ctry in c("Germany", "India", "Bangladesh")) {
  cat(sprintf(
    "%s: %d reported dates with a gap of more than 1 day before interpolation\n",
    ctry, n_gaps(vaccinations_country, ctry, "people_fully_vaccinated_per_hundred")
  ))
}
```

---

## Germany: ARIMA Model

```{r germany_ts}
germany_series <- prepare_country_series(
  vaccinations_country, "Germany", "people_fully_vaccinated_per_hundred"
)

ts_germany <- ts(germany_series$value, frequency = 7)
```

The cumulative coverage series is, by construction, monotonically
non-decreasing, so it is not stationary in levels. Rather than assuming a
fixed order of differencing, `auto.arima()` is left to select `d` and `D`
itself (via its internal KPSS/OCSB unit-root tests), which is a more
defensible default than hard-coding `d = 1, D = 1` up front.

```{r germany_arima, message=FALSE, warning=FALSE}
fit_germany <- auto.arima(ts_germany, stepwise = FALSE, approximation = FALSE)
summary(fit_germany)
cat("\nSelected order (p, d, q)(P, D, Q)[m]:\n")
print(arimaorder(fit_germany))
```

```{r plot_forecast_function}
plot_forecast_teal <- function(fc, title_text, y_label) {
  fitted_vals <- as.numeric(fc$x)
  forecast_vals <- as.numeric(fc$mean)

  n_fitted <- length(fitted_vals)
  n_fore   <- length(forecast_vals)

  fitted_df <- data.frame(t = seq_len(n_fitted), value = fitted_vals)

  fore_df <- data.frame(
    t    = n_fitted + seq_len(n_fore),
    mean = forecast_vals,
    lo80 = as.numeric(fc$lower[, 1]),
    hi80 = as.numeric(fc$upper[, 1]),
    lo95 = as.numeric(fc$lower[, 2]),
    hi95 = as.numeric(fc$upper[, 2])
  )

  ggplot() +
    geom_ribbon(data = fore_df, aes(x = t, ymin = lo95, ymax = hi95),
                fill = teal_pale, alpha = 0.6) +
    geom_ribbon(data = fore_df, aes(x = t, ymin = lo80, ymax = hi80),
                fill = teal_light, alpha = 0.7) +
    geom_line(data = fitted_df, aes(x = t, y = value),
              colour = "#555555", linewidth = 0.7) +
    geom_line(data = fore_df, aes(x = t, y = mean),
              colour = teal_base, linewidth = 1) +
    labs(
      title    = title_text,
      subtitle = "30-day forecast with 80% and 95% confidence intervals",
      x        = "Time (days)",
      y        = y_label
    ) +
    theme_teal()
}
```

```{r germany_forecast}
forecast_30_germany <- forecast(fit_germany, h = 30)

plot_forecast_teal(
  forecast_30_germany,
  title_text = "ARIMA Forecast: Germany Full Vaccination Coverage",
  y_label    = "Fully vaccinated per 100"
)
```

---

## Residual Diagnostics

```{r residuals}
checkresiduals(fit_germany)
```

The residual diagnostics above should be read directly from your own knit
output rather than assumed: check that (1) residuals fluctuate around zero
with no obvious trend, (2) the ACF shows no strong remaining autocorrelation
beyond the seasonal lag (7, 14, ...), and (3) the Ljung-Box p-value printed
by `checkresiduals()` is not significant at your chosen alpha. If any of
these fail, increase `max.p`/`max.q` in `auto.arima()` or reconsider the
differencing order before trusting the forecast intervals.

---

## ETS Model Comparison

```{r ets_forecast}
fit_ets_germany <- ets(ts_germany)
fc_ets_germany  <- forecast(fit_ets_germany, h = 30)

plot_forecast_teal(
  fc_ets_germany,
  title_text = "ETS Forecast: Germany Full Vaccination Coverage",
  y_label    = "Fully vaccinated per 100"
)
```

---

## ARIMA vs ETS Accuracy (Train/Test Split)

```{r accuracy_comparison, message=FALSE, warning=FALSE}
n_obs  <- length(ts_germany)
n_train <- round(n_obs * 0.8)

train_vals <- as.numeric(ts_germany)[1:n_train]
test_vals  <- as.numeric(ts_germany)[(n_train + 1):n_obs]

ts_train <- ts(train_vals, frequency = 7)

fit_arima_train <- auto.arima(ts_train, stepwise = FALSE, approximation = FALSE)
fit_ets_train   <- ets(ts_train)

h <- length(test_vals)

acc_arima <- accuracy(forecast(fit_arima_train, h = h), test_vals)
acc_ets   <- accuracy(forecast(fit_ets_train,   h = h), test_vals)

cat("ARIMA test-set accuracy:\n"); print(acc_arima)
cat("\nETS test-set accuracy:\n");   print(acc_ets)
```

Compare the `RMSE`/`MAE` rows for the **Test set** between the two tables
above (not the training-set rows) to decide which model generalises better
for Germany; do not assume ARIMA is automatically the stronger model.

---

## Multi-Country ARIMA Forecast (Inequality Comparison)

Each of the three countries below is modelled the same way as Germany
above — `auto.arima()` on the gap-filled **level** series, letting the
function choose its own differencing order. This deliberately replaces the
previous version's approach, which manually first-differenced the series
into `daily_change` *and then* still forced `d = 1, D = 1` inside
`auto.arima()` — a double-differencing bug that over-differences an
already-stationary series and destabilises the fitted model. Using one
consistent modelling strategy across all countries also makes the
cross-country forecast-uncertainty comparison below fair (like for like).

```{r multicountry_arima, message=FALSE, warning=FALSE}
countries_arima <- c("Spain", "India", "Bangladesh")
fits <- list()
forecasts <- list()

for (ctry in countries_arima) {

  series <- prepare_country_series(
    vaccinations_country, ctry, "people_fully_vaccinated_per_hundred"
  )

  if (nrow(series) < 21) {
    cat(sprintf("Skipping %s — fewer than 21 usable observations (n = %d)\n",
                ctry, nrow(series)))
    next
  }

  ts_ctry <- ts(series$value, frequency = 7)

  adf_p <- tryCatch(
    suppressWarnings(tseries::adf.test(ts_ctry)$p.value),
    error = function(e) NA_real_
  )

  fit_ctry <- auto.arima(ts_ctry, stepwise = FALSE, approximation = FALSE)
  fc_ctry  <- forecast(fit_ctry, h = 30)

  fits[[ctry]] <- fit_ctry
  forecasts[[ctry]] <- fc_ctry

  cat(sprintf(
    "\n%s: n = %d, ADF p-value (levels) = %s, selected order = %s\n",
    ctry, nrow(series),
    ifelse(is.na(adf_p), "NA", sprintf("%.3f", adf_p)),
    paste(arimaorder(fit_ctry), collapse = ",")
  ))

  print(
    plot_forecast_teal(
      fc_ctry,
      title_text = paste("ARIMA Forecast of Full Vaccination Coverage:", ctry),
      y_label    = "Fully vaccinated per 100"
    )
  )
}
```

The interpretation of these forecasts should be read off your own knit
output rather than assumed in advance. Spain, India, and Bangladesh differ
substantially in how much differencing `auto.arima()` selected and in
whether the ADF test rejects a unit root in levels — report the actual
`arimaorder()` output and p-value for each country here, rather than a
generic expectation, since (as the next section shows) the naive
"richer country → more reliable forecast" assumption does not hold up
once it is actually measured.

---

## Quantifying Forecast Uncertainty as an Inequality Metric

Rather than comparing the three forecasts only by eye, this section
computes a single number per country — the mean width of the 95%
prediction interval across the 30-day forecast horizon — and treats it as
a quantitative indicator of *data-driven planning uncertainty*. This ties
the ARIMA analysis directly back to the report's research question: it is
one thing to say lower-income countries vaccinated more slowly, and
another to show that the reliability of any forecast made from their own
reported data is itself unequal.

```{r uncertainty_table}
uncertainty_summary <- lapply(names(forecasts), function(ctry) {
  fc <- forecasts[[ctry]]
  width_95 <- as.numeric(fc$upper[, 2]) - as.numeric(fc$lower[, 2])
  mean_level <- mean(as.numeric(fc$mean))
  data.frame(
    country            = ctry,
    n_observations      = length(fc$x),
    mean_pi_width_95    = mean(width_95),
    max_pi_width_95     = max(width_95),
    # width scaled by the forecast's own average level, so a country
    # forecast to sit near 70% coverage isn't unfairly penalised for
    # having more absolute room to swing in than one still near 1%.
    relative_pi_width_95 = mean(width_95) / mean_level
  )
}) %>% bind_rows() %>% arrange(desc(mean_pi_width_95))

print(uncertainty_summary)

ggplot(uncertainty_summary,
       aes(x = reorder(country, mean_pi_width_95), y = mean_pi_width_95)) +
  geom_col(fill = teal_base, width = 0.6, colour = "white") +
  geom_text(aes(label = sprintf("%.1f", mean_pi_width_95)),
            hjust = -0.2, colour = teal_dark, fontface = "bold") +
  coord_flip() +
  expand_limits(y = max(uncertainty_summary$mean_pi_width_95) * 1.15) +
  labs(
    title    = "Average Width of the 95% Forecast Interval by Country",
    subtitle = "Mean over a 30-day horizon, in percentage points — absolute scale",
    x        = NULL,
    y        = "Mean 95% PI width (percentage points)"
  ) +
  theme_teal()

ggplot(uncertainty_summary,
       aes(x = reorder(country, relative_pi_width_95), y = relative_pi_width_95)) +
  geom_col(fill = red_mid, width = 0.6, colour = "white") +
  geom_text(aes(label = scales::percent(relative_pi_width_95, accuracy = 0.1)),
            hjust = -0.2, colour = red_dark, fontface = "bold") +
  coord_flip() +
  expand_limits(y = max(uncertainty_summary$relative_pi_width_95) * 1.15) +
  labs(
    title    = "Forecast Interval Width Relative to Forecast Level",
    subtitle = "95% PI width as a share of the country's own forecast mean — scale-adjusted",
    x        = NULL,
    y        = "Relative 95% PI width"
  ) +
  theme_teal()
```

**Read the two charts above together, not just the first one.** In this
run, Spain's *absolute* 95% interval was the widest of the three (about
7.0 percentage points, versus roughly 2.0 for India and 1.4 for
Bangladesh) — the opposite of the naive "richer country, more reliable
data, narrower interval" story this section set out to test. That is
worth reporting honestly rather than reinterpreted to fit the
expectation. A plausible explanation, visible directly in the fitted
orders: Spain's model needed both non-seasonal *and* seasonal
differencing (its `arimaorder()` shows `D = 1`), while India's and
Bangladesh's models used only non-seasonal differencing — seasonal
differencing compounds uncertainty faster over a 30-day horizon. Spain
was also already close to saturation (around 70 fully vaccinated per
100) by the end of the series, where absolute week-to-week swings have
more room to occur than in a series still crawling upward from near
zero — which is exactly why the *relative* chart (interval width as a
share of the forecast's own level) is shown alongside the absolute one:
if Spain's relative ranking drops below India's and Bangladesh's once
rescaled, that supports "more absolute room to move, not worse
underlying data," whereas if it stays highest even in relative terms,
that's stronger evidence of a genuine reliability gap. Whichever pattern
your own numbers show, report the actual ranking from both charts, not
the assumption stated at the start of this section.

---

## Summary

This notebook fitted ARIMA models to gap-filled daily vaccination coverage
series for Germany (single-country deep dive), and Spain, India, and
Bangladesh (multi-country comparison), using one consistent methodology
(auto-selected differencing on levels) throughout, and converted the
resulting forecast uncertainty into two comparable metrics per country —
absolute and relative 95% interval width. The residual diagnostics and
accuracy comparison for Germany provide a validity check for the
modelling approach before it is extended to the other countries. The
headline result of the comparison is not the simple "poorer country,
less reliable forecast" pattern originally hypothesised — Spain's
absolute forecast uncertainty came out highest, not lowest — and that
result, along with its most likely explanation (near-saturation
volatility and seasonal differencing, not necessarily worse underlying
data), should be reported as found rather than adjusted to match the
initial expectation.